---
title: "Cox regression"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

In this tutorial, we will do a basic demonstration of how to run a Cox regression 
in R. 

Remember from lecture that Cox regression models the hazard curve, which is 
defined as the instantaneous risk of death (or having the event) given that one 
hasn't had the event up that point. This is a bit abstract! But a couple things 
to note: 

* Like the survival curve, the hazard curve is a function over time. 
* Higher hazards at a certain time mean higher risk of death at that timepoint; 
lower hazards mean lower risk of death. 

In Cox regression, the output is a **hazard ratio.** The hazard ratio describes 
how much hazard curves differ between different groups. For example, a hazard ratio
of 5 signifies that the hazard of death of one group is 5 times higher than that 
of another group. 

## Example

We will run an Cox example regression in R. Again, we will use the `survival` package 
and the lung dataset. We start off with some basic data processing and creating a
time to event variable with the `Surv()` function: 

```{r, warning = F, message = F}
library(survival)
library(dplyr)

# convert status variable to 0s and 1s 
# convert sex into factor
lung_new <- lung %>% 
  mutate(status = ifelse(status == 2, 1, 0), 
         sex = factor(sex))

# create time to event variable with Surv() function
lung_new <- lung_new %>% 
  mutate(tte = Surv(time, status))

```

Suppose that we want to perform a Cox regression to compare the hazard of death
between females and males. 

To run the regression, we use the `coxph()` function from the `survival` package. 
The first argument of this function is the formula, which is similar to 
linear/logistic regression: `outcome ~ predictor`. In this case, the outcome is
the time to event (`tte`) and the predictor is sex (`sex`). The second 
argument is the dataframe. 

```{r}
cox_mod <- coxph(tte ~ sex, data = lung_new)

summary(cox_mod)
```
In this output, we will focus on the exponentiated output, specifically exp(coef), 
lower .95, and upper .95, as well as the p-value (Pr (|z|))

* The exponentiated outputs give us an estimate and confidence interval of the 
hazard ratio comparing sex = 2 (females) to sex = 1 (males). We see a hazard 
ratio of 0.588, so the hazard of death for females is 0.588 times that of males. 
In other words, the hazard of death for females is 41% lower than that of males. 
* The p-value tests the hypothesis that the hazard curves are the same. In this case, 
we see a p-value of 0.001, so we reject the null hypothesis that the two 
hazard curves are the same. 

## Exercise

Run a Cox proportional hazards model, regressing the time-to-event on the age. 
Provide an interpretation on the coefficient of age. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

