---
title: "Creating a Table 1"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
```

For assignment 3, your goal is to create a "Table 1" with descriptive statistics 
for relevant variables using R. 

The goal is to create a table which describes baseline characteristics stratified 
by treatment arm. Your table should look something like this:

```{r, echo = FALSE}
# create Table 1 table structure
data.frame(
  Variable = c("", "Age", "Female", "etc"),
  `BUP-NX Treatment` = c("Mean (sd) or n (%)", "", "", ""),
  `XR_NTX Treatment` = c("Mean (sd) or n (%)", "", "", "")
) %>% 
  kable()
```

You should be able to do this based on what we have learned in R so far. An alternative
option - look into the `table1` package online. This will make things much faster. 
But make sure to convert all categorical variables into factor variables before
using this package. 

Good luck!

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*



