---
title: "Factor variables"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Variable types

Each dataframe column can have different types. Some variable are *string* variables 
(e.g. text); others are *numeric* variables (e.g. numbers). There is another type 
of variable in R, called a **factor**. These are categorical variables, which often 
are coded as integers in a dataset (e.g. 0, 1, 2, etc.). 

One example: we may want to consider the variable `tx` in the crisp-2024-sample100.csv 
dataset as a factor variable because it only takes two distinct values: 0 and 1, where
0 corresponds to BUP-NX and 1 corresponds to XR-NTX. 

```{r}
df <- read.csv("crisp-2024-sample100.csv")

table(df$tx)
```

By default, R will read in this variables as in *numeric*. It starts off by thinking this column is just a number. But, we may want to change 
it into a factor before we perform our analysis. To do this, we can use the following line of code: 

```{r}
df$tx <- factor(df$tx)
```

This code to the right hand side of `<-`  converts the `tx` column of `df` 
into a factor, and then we save it back into the `tx` column of `df`. 

We can tell that it is a factor because "Levels: 0 1" is displayed on the bottom of the output when printing `df$tx`.

```{r}
df$tx
```

We can also provide distinct labels to this variable. To do this, we specify 
an additional argument: `labels = c("BUP-NX", "XR-NTX")` in the output. This will 
assign the 0 to BUP-NX and 1 to XR-NTX. 

```{r}
df$tx <- factor(df$tx, labels = c("BUP-NX", "XR-NTX"))

df$tx
```

It is generally a good idea to convert categorical variables into factors
before performing analysis in R. That way, R knows that these are categorical variables 
and not numbers. 

## Exercises 

Examine the data dictionary for crisp-2024-sample100.csv. Which are categorical variables? Write some code to convert these columns into factors. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*


