---
title: "Functions"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Walkthrough

In this tutorial, we will learn how to write our own functions. We've used many 
pre-built functions before - for example, `mean()`, `sd()`, `summary()`, 
`table()`, etc. But you may want to write your own function if it's a task 
that you want to perform repeatedly. This can save you time writing code, and can 
also make the code more readable and reproducible! 

To write a function, we take the following steps: 

1. Specify the function name. In the code below, we call our function `divide_by_two`. 
2. Next, assign it to a `function()` object, where the **arguments** (i.e. inputs) go 
inside the parenthesis. In the example below, we take as an argument one number `my_num`. 
3. After that, we write out what we want the function to do inside the brackets
\{ \}. *The last line of the function will be the output of the function.* For example, 
the function below takes the input my_num and divides it by 2. 

```{r}
divide_by_two <- function(my_num){
  my_num / 2
}
```

We can apply this function on different inputs: 

```{r}
divide_by_two(1)

divide_by_two(100)
```

Optionally, we can also include the argument name `my_num = ` when calling the 
function. 

```{r}
divide_by_two(my_num = 1)

divide_by_two(my_num = 100)
```

Let's write a more complicated function that might be more relevant for data 
analysis. Suppose that we want a function that takes in as input a dataframe, and 
outputs a **list**, which has stored the standard deviation of each column. We can do that
using the code below: 

```{r}
find_sds <- function(df_input){
  
  # pull the column names as a vector
  cols <- names(df_input)
  
  # initialize and empty list 
  output <- list()
  
  # for each column, find the standard deviation of each column 
  for (col in cols){
    output[[col]] <- sd(df_input[,col], na.rm = T)
  }
  
  # return the output 
  output
}
```

This code is a bit complicated, so let's walk through it. First, our function 
name is `find_sds`, and it takes in one argument `df_input`. Inside our function, 
we do the following: 

1. Pull out the column names and save it as the `cols` variable. 
2. Create an empty list called `output`. 
3. Loop through all of the columns in the `cols` variable: 
  * For each column, save the standard deviation of the column with their corresponding
  label inside the `output` list. 
4. Output the `output` variable. 

Let's apply this function to the X:BOT data. 

```{r}
df <- read.csv("crisp-2024-sample100.csv")
```

```{r}
find_sds(df)
```

## Exercise 

Write a function which takes as an input a dataframe and outputs a list which includes the number of cells with `NA` for each column.

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

