---
title: "Graphing with ggplot2"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

In this tutorial, we will learn the basics of `ggplot2`, which is a very popular 
plotting package. R contains default plotting functions (with the `plot()` function), 
but these are often not intuitive and difficult to customize. `ggplot2` provides a more
intuitive and customizable interface for graphing data. 

For this tutorial, we will use the same dataset as before. 

```{r, warning = F, message = F}
# read dplyr library 
library(dplyr)

# read dataset 
df <- read.csv("crisp-2024-sample100.csv")

# convert some columns into factors
df$tx <- factor(df$tx, 
                labels = c("BUP-NX", "XR-NTX"))

df$sex <- factor(df$sex, 
                 labels = c("Male", "Female"))

```

If you haven't yet, install the `ggplot2` function and call in into your R session.

```{r, warning = F}
# install.packages("ggplot2")

library(ggplot2)
```


## ggplot2 syntax

To create a `ggplot2` graph, you start off with the `ggplot()` function. Within
the `ggplot()` function, you specify the dataset and which columns you want to put 
in each axis in an aesthetic layer `aes(x = x_var, y = y_var)`. 

```{r}
ggplot(data = df, aes(x = age, y = bmi))
```

In our `aes()` call, we want `age` to be on our x-axis, and `bmi` to be on our y-axis. 
Note that this code plots an empty graph. That is because we haven't specified 
what objects we want to include in our plot. To do this, we use `+` to add geometric features 
to our plot. 

## Scatter plots 

If we want to create a scatter plot, we can add `+ geom_point()` to our 
code. For example: 

```{r, warning = F}
ggplot(data = df, aes(x = age, y = bmi)) + 
  geom_point()
```

Now we have successfully graphed a scatter plot! Let's now make some customizations. 

Suppose that we want the colors of the points to correspond to the treatment 
arm. We can add a `color = tx` argument within the `aes()` call: 

```{r, warning = F}
ggplot(data = df, aes(x = age, y = bmi, color = tx)) + 
  geom_point()
```

We can also change the "look" of our plot. For example, if we add `+ xlim(0, 60)` and `+ ylim(0,40)`,
we can change the x and y axis ranges of our plot to 0-60 and 0-40, respectively. 

```{r, warning = F}
ggplot(data = df, aes(x = age, y = bmi, color = tx)) + 
  geom_point() + 
  xlim(0,60) + 
  ylim(0,40)
```

We can also add a title to our graph using `ggtitle()`: 

```{r, warning = F}
ggplot(data = df, aes(x = age, y = bmi, color = tx)) + 
  geom_point() + 
  xlim(0,60) + 
  ylim(0,40) + 
  ggtitle("My first graph")
```

We can continue to add customizations using `+`. Learning all the customizations will take time. 
In the meantime, you can easily Google (or ask ChatGPT) about customizations. For example, 
you can look up "How do I change the x-axis label in ggplot?" in Google, and it will tell you how 
to change that configuration. 

## Bar plots 

We can also create bar plots using `+ geom_bar()`. 
For example, suppose that we wanted to graph the number in each treatment arm.
We can write the following: 

```{r}
ggplot(data = df, aes(x = tx)) + 
  geom_bar()
```

Note that we don't need a `y = y_var` specification in our `aes()` call. This
is because `geom_bar()` by default counts the number in each category and uses 
that number as the height of the bar. So no need to specify the y-axis of this graph. 

Now, suppose that we want to stratify the bars by sex. We can do this by adding 
a `fill` to our `aes()` call: 

```{r}
ggplot(data = df, aes(x = tx, fill = sex)) + 
  geom_bar()
```

We now have different colors indicating what proportion of each bar represent
males and females. 

## Histograms 

We can create histograms by `+ geom_histogram()`. For example, suppose
that we want to create a histogram to look at the distribution of ages in our dataset. 

```{r, message = F}
ggplot(data = df, aes(x = age)) + 
  geom_histogram()
```

We can customize the histogram by including additional arguments within the 
`geom_histogram()` function call. For example, to decrease to 5 bins (which increases the
width of the bins), we can write the following:

```{r, message = F}
ggplot(data = df, aes(x = age)) + 
  geom_histogram(bins = 5)
```

## Boxplots 

Another common type of plot is a boxplot. In this type of plot, we can visually 
compare the distributions of a variable between different categories. For example, suppose that we
wanted to compare the ages between the treatment arms. We can create a boxplot using 
`+ geom_boxplot()`: 

```{r}
ggplot(data = df, aes(x = tx, y = age)) + 
  geom_boxplot()
```

## Exercise

1. Create a graph with two bars (males and females) with each bar consisting of two colors, where
each color represents the count of participants with and without lifetime usage of cocaine at enrollment. 

2. Create a scatter plot of age versus BMI. 

3. Create a boxplot comparing the distribution of BMIs between males and females. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

