---
title: "Installing and using packages"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

One of the main advantages of R are its packages. 

An R package is a collection of functions, data, and documentation that extends the capabilities 
of R, a programming language and software environment used for statistical computing 
and graphics. R packages are used to share code and data with others and to 
reuse code in different projects. They can include:

1. **Functions**: Custom functions that perform specific tasks.
2. **Data**: Datasets that can be used for analysis and examples.
3. **Documentation**: Help files and vignettes that explain how to use the functions and datasets.

R packages are typically installed from CRAN (Comprehensive R Archive Network). Packages
tend to have specific uses. For example, there are packages specifically for 
making graphs, others for creating manuscript-ready tables, and others to simplify 
data processing. 

You can install a package in R using the `install.packages()` function, and load 
it using the `library()`.

```{r eval = F}
# Install a package from CRAN
install.packages("dplyr")

# Load the package
library(dplyr)
```

Most packages have documentation on how to use specific functions. For example, 
the `dplyr` documentation is [here](https://dplyr.tidyverse.org/). 

Each time you use R, you will have to load in the packages that you want to use
during the session. 

## Exercises

Install and load in the following packages: `dplyr` and `ggplot2`.

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

