---
title: "Linear regression with an interaction term"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

In this tutorial, we will do one full example of a linear regression with an 
**interaction term**. We will practice interpreting the output from running this
linear regression in R. 

```{r}
df <- read.csv("crisp-2024-sample100.csv")
```

## Example 

Suppose that wish to fit a linear regression, regressing BMI on age, sex, and we 
want to include an interaction term between age and sex. We may wish to do this if we
believe that the association between age and BMI differ for males and females. 

The model we are trying to fit looks like: 

$$
bmi = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 age * sex
$$

Note that within this linear regression, we include *main effect* terms for each 
age and sex ($\beta_1 age$ and $\beta_2 sex$), as well as an *interaction term* 
($\beta_3 age*sex$). 

To find the values of $\beta_0, \beta_1, \beta_2, \beta_3$ that best fit the data, 
we can again use the `lm()` function in R. To code an interaction term, you can 
use the syntax `outcome ~ pred1 * pred2` in the formula. This will automatically 
create main effect terms for `pred1` and `pred2`. In this example, we can code
the formula as `bmi ~ age * sex`. 

```{r}
mod <- lm(bmi ~ age * sex, data = df)

summary(mod)
```
As you can see, we now have the estimates for the intercept, 
the coefficient on age, the coefficient on sex, and the coefficient on the interaction 
between age and sex. 

How can we interpret this output? 

* *Intercept*: The predicted BMI for males with age = 0. Note that this is not 
meaningful, since we don't have age = 0 in our population of interest.
* *Coefficient on age*: The estimated difference in BMI comparing two groups **of males** 
differing in one year of age. Note that this interpretation is different that of a 
normal adjustment model. 
* *Coefficient on sex*: The estimated difference in BMI comparing groups of females versus 
males **with age = 0**. Note that again this interpretation is different that of a 
normal adjustment model. 
* *Coefficient on age:sex*: The estimated difference in the mean difference in 
BMI for every 1-year difference in age in females versus that in males. This is 
a tricky interpretation ("a difference of differences")! Essentially, this 
coefficient tells us how different the association between BMI and age is 
between males and females. 
  * *p-value*: The result from testing the null hypothesis that the coefficient on 
  the interaction coefficient is 0. With $p = 0.33$, we fail to find evidence of an 
  interaction effect between age and sex in their association with BMI. 


## Exercise

Perform a linear regression, with *BMI as the outcome* and *cannabis use history and cocaine use history as predictors*. 
Include an interaction term between the two predictors. Provide an interpretation of the output. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

