---
title: "Lists"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

There are several different types of **data structures** in R. We've learned 
about a few of them: 

* **Numeric**: a number
* **String**: a text snippet
* **Factor**: a categorical variable
* **Vector**: a ordered sequence of multiple numbers or strings
* **Dataframe**: a data table with rows and columns 

```{r}
# numeric
my_num <- 1

# string 
my_str <- "This is my string"

# vector
my_vec <- c(1,2,3,4)
my_vec2 <- c("yes", "no")

# the cars object (included in R) is a dataframe
head(cars)
```


Another data structure that is often useful is a **list**. A list is a *labeled* object 
that can consist of elements of different types. To create a list, you can use the 
syntax `list(label1 = obj1, label2 = obj2, ...)` where the label1, label2, etc.
represent the labels for each object. 

For example: 

```{r}
my_list <- list(
  name = "John Doe",
  age = 30,
  scores = c(90, 85, 88)
)
```


In this list, we have three objects: 

1. A string "John Doe" with the label "name"
2. A numeric 30 with the label "age" 
3. A vector `c(90, 85, 88)` with the label "scores"

We can pull out different objects by calling the list with a $, along with the label 
for the object of interest. For example, if 
we want to pull out the name and age from `my_list`, we can do the following: 

```{r}
# pull out name
my_list$name

# pull out age 
my_list$age
```


A list is a very flexible and versatile object to store different types of data, 
along with clear labels. They are commonly used in the output of functions, 
which we will talk about in a future tutorial. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

