---
title: "Logistic regression"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

In this tutorial, we will do one full example of a univariate (one predictor)
logistic regression, and we will learn how to interpret the regression 
coefficients of the output.

```{r}
df <- read.csv("crisp-2024-sample100.csv")
```

## Example 

In logistic regression, the outcome is binary - either 0 or 1. For example, 
in this example, suppose that we consider history of psychotic diagnosis (`mhdx`) 
as our outcome, and age as our exposure. 

In the logistic model, we model the **odds** of the outcome. In this example, we 
are modeling the odds of having a history of psychotic diagnosis, which we
denote as $Odds(mhdx)$. Logistic regression also uses a logarithmic transform on 
this outcome, for reasons not explained in this tutorial. So our model is:  

$$
log(Odds(mhdx)) = \beta_0 + \beta_1 age
$$

The right hand side of the model should be familiar; this is a just a normal linear
model. 

Equivalently, we can exponentiate both sides and write this as: 

$$
Odds(mhdx) = exp(\beta_0 + \beta_1 age) 
$$

To run this in R, we can use the `glm()` function. `glm` stands for "generalized 
linear model," which is a more flexible form of a linear model. Within the `glm()` 
function call, we specify the: 

1. Formula: `outcome ~ predictor` (similar to linear regression)
2. Data: `data = df` 
3. Family: `family = binomial` (this configures the function to run logistic regression)

So our function call looks like this: 

```{r}
mod <- glm(mhdx ~ age, data = df, family = binomial)

summary(mod)
```
The estimates for the intercepts and age correspond to estimates for $\beta_0$ and
$\beta_1$, respectively. In logistic regression, it is often more useful to interpret 
the exponentiated coefficient estimates. To do that, we exponentiate the 
coefficient table using the `exp()` function on the `coef()` table from 
the `mod` object: 

```{r}
exp(coef(mod))
```

How can we interpret these coefficients?

* *Intercept*: The exponentiated intercept represents the predicted odds of having a 
history of psychotic diagnosis for someone with age = 0. That is, we estimate that 
the odds of having a history of psychotic diagnosis for someone with age = 0 is 0.75. 
* *Coefficient on age*: The exponentiated coefficient on age represents the odds ratio 
comparing the odds of having a history of psychotic diagnosis for two groups differing 
by one year of age. That is, we estimate that for two groups differing in one year of age, 
the older group has around a 4% lower odds ($1 - 0.962$) of having a history of psychotic
diagnosis compared to the younger group. 
  * *p-value*: The result from testing the null hypothesis that the coefficient on 
  age is 0 (or equivalently the exponentiated coefficient is 1). With $p = 0.275$, we 
  fail to find a significant association between age and history of psychotic diagnosis. 

To obtain confidence intervals for these exponentiated estimates, we can 
exponentiate the output of `confint(mod)`: 

```{r, message = F}
exp(confint(mod))
```


## Exercise

Perform logistic regression with cannabis use as the outcome and sex as the predictor. 
Provide an intepretation of the exponentiated coefficient estimates. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

