---
title: "Loops"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Walkthrough

In this tutorial, we will learn about how to run a loop in R. This is useful if
we want to perform the same operation multiple times. 

For example, suppose that we want to add 2 to each number in a vector which 
consists of c(1,2,3,4,5). We can create a loop to help do this. 

1. The loop starts with a `for` statement. 
2. After the `for` statement, we include inside the parenthesis () the instructions 
for the loop. The instructions consist of "___ in ___". In the first blank, 
we provide a temporary label for the object in the loop, so that we can reference it later on. In the 
second blank, we put the name of the vector we want to loop through. 
3. Finally, we include instructions for each iteration of the loop inside the brackets 
\{ \}. 

Here is a toy example: 

```{r}
my_vec <- c(1,2,3,4,5)

for (num in my_vec){
  print(num + 2)
}
```

In this example, we are looping through `my_vec`. We temporarily name objects that 
we loop through as `num`. And for each iteration of the loop, as we go through 
1,2,3,4,5 in `my_vec`, we print out each number `num` PLUS 2. 

Here is another example, which may be more relevant in data analysis. Suppose 
that we wanted to create a loop to print out the *mean* for each column in a dataframe. 

Let's use the X:BOT dataset, which we have used for previous tutorials. 

```{r}
df <- read.csv("crisp-2024-sample100.csv")

head(df)
```

Let's create a loop to find the mean for each column. We want to loop through all 
the columns of `df`. To pull out all the columns as a vector, you can use the 
`names()` function. 

```{r}
# pull out the column names of the dataframe
cols <- names(df)

cols
```

Now let's write our loop. We start with `for ()`, and inside the parenthesis 
we include `col in cols`. This says that we want to loop through the columns (which 
we saved as the variable `cols`), and as we do that, we temporarily label each column as `col`. 

Finally, within the brackets \{ \}, we print out the name of the column and the 
mean for the column in the dataframe.


```{r}
for (col in cols){
  print(col)
  print(mean(df[,col], na.rm = T))
}
```
Loops are very powerful, so that you don't have to rewrite the same code over and over 
again!

## Exercise 

Loop through all the binary variables in the X:BOT dataset and print out a tabulation of 
the number in each category. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

