---
title: "Reading in data into R"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Say that you have a dataset. The first step is to read in data into R. You can do
this using the `read.csv()` function. 

The easiest way to use this function is to put the data file in the **same folder** 
as your R (or Rmd) code. For example, please download the crisp-2024-sample100.csv 
file from ITHS Tree and put it in the same folder as this R Markdown file, 
and try to run the following code: 

```{r}
# we read the crisp-2024-sample100.csv and save it into df 
df <- read.csv("crisp-2024-sample100.csv")

# use head function to print out the first few rows
head(df)
```

Did it run? Then great! That means that the data is now in our R session, saved
in the variable `df`. We can now manipulate and analyze the data, which we will 
do in the next few tutorials.

If the data is not in the same folder as your R file, you may have to include the directory path to the data in the argument when calling `read.csv()`.

```{r}
# if data is in a different folder, it may look like this: 
# df <- read.csv("C:/path/to/directory/crisp-2024-sample100.csv")
```

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*
