---
title: "Risk difference, risk ratios, and odds ratios"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

The purpose of this tutorial is to understand the difference between risk differences, 
risk ratios, and odds ratios. We will calculate these by hand using R. 

For this tutorial, we will use the dataset that we have been using: 

```{r, warning = F, message = F}
library(dplyr)

df <- read.csv("crisp-2024-sample100.csv")
```

Suppose our research question is the following: 
*What is the difference in the proportion of lifetime use of cocaine between males and females?*

## Risk difference

The first way to quantify this difference is by calculating the risk difference. 
The risk difference is the difference in the proportion who had lifetime cocaine usage between 
males and females. 

$$
RD = p_{females} - p_{males}
$$

where $p_{males}$ is the proportion of cocaine usage among males, and $p_{females}$ is 
the proportion among females. 

To calculate this in R, we first create a 2x2 table: 

```{r}
sex_coc_tbl <- table(df %>% select(sex, coclft))

sex_coc_tbl
```

The proportion of males who used cocaine is $\frac{38}{33 + 38} = 0.535$.

```{r}
38 / (33 + 38)
```

The proportion of females who used cocaine is $\frac{15}{15 + 14} = 0.517$.

```{r}
15 / (15 + 14)
```

So the risk difference is the difference between these two quantities: 

```{r}
0.517 - 0.535
```

This tells us that estimated *raw difference* in the percentage of cocaine usage between 
the two groups is 1.8%, with females having the lower percentage of usage. 

## Risk ratio 

Another way of quantifying the difference is by using the risk ratio. This equals: 

$$
RR = \frac{p_{females}}{p_{males}}
$$ 

In our example, we obtain: 

```{r}
0.517 / 0.535 
```

This tells us the *relative difference* in cocaine usage between the groups. 
That is, females have an estimated 3.4% lower risk cocaine usage compared
to males. 

## Odds ratio

The last way of quantifying the the difference is with the odds ratio. The 
odds ratio equals the following: 

$$
OR = \frac{odds_{female}}{odds_{male}}
$$

where $odds_{male}$ is the *odds* of cocaine usage among males, and $odds_{female}$ is 
the *odds* among females. 

Odds is not the same as probability! Let's take a look at one example of the difference. 
Here again is our 2x2 table: 

```{r}
sex_coc_tbl
```

Among males (`sex == 0`), the proportion who used cocaine is $38/(33+38)$ whereas
the odds of cocaine usage is $38/33$. 

```{r}
# proportion of cocaine usage among males
38 / (33 + 38)

# odds of cocaine usage among males
38 / 33
```

The odds of females who used cocaine in our dataset is: 

```{r}
15 / 14
```

Therefore, we calculate the odds ratio to be: 

```{r}
1.071 / 1.152 
```

Therefore, we estimate a 7% lower odds of cocaine usage among females compared 
to males.

## Confidence intervals

We can calculate these estimates by hand, but we may want confidence intervals to 
obtain some measure of uncertainty around our estimate. We can use the `epitools` 
package to calculate confidence intervals for the risk ratio and the odds ratio.

```{r}
# install.packages("epitools")

library(epitools)
```

To find the confidence interval for the *risk ratio*, we can use the function 
`riskratio.wald()`, with the 2x2 table as our argument

```{r}
riskratio.wald(sex_coc_tbl)
```

We look at the `$measure` portion of the output. We see that the risk ratio is 
0.966 (same as what we calculated by hand), with 95% CI of 0.639 - 1.461. 

To find the confidence interval for the *odds ratio*, we can use the function 
`oddsratio.wald()`, with the 2x2 table as our argument: 

```{r}
oddsratio.wald(sex_coc_tbl)
```

Again, we look at the `$measure` portion of the output. We see that the odds ratio is 
0.930 (same as what we calculated by hand), with 95% CI of 0.392 - 2.210.

We can also calculate a confidence interval on the risk difference. There are several 
packages/functions to do this (e.g. take a look at the`riskdifference()` function 
from the `fmsb` package).


## By subgroups

You may want to calculate each of these measures for different subgroups, in order
to check for confounding. Doing this is straightforward using what we have learned so
far. For example, suppose that we want to calculate the risk ratios for each treatment
arm.  

We first create dataframes that filter for each treatment arm: 

```{r}
df_tx0 <- df %>% filter(tx == 0)
df_tx1 <- df %>% filter(tx == 1)
```

We then create 2x2 tables of sex and cocaine usage for each treatment arm. 

```{r}
table_tx0 <- table(df_tx0 %>% select(sex, coclft))
table_tx1<- table(df_tx1 %>% select(sex, coclft))
```

Finally, we can use the `riskratio.wald()` function to calculate estimates and 
confidence intervals for the risk ratios for each group: 

```{r}
riskratio.wald(table_tx0) 
```

```{r}
riskratio.wald(table_tx1)
```


## Exercise

Calculate the risk difference, risk ratio, and odds ratio of lifetime marijuana
usage comparing males and females in the dataset. 

Calculate the odds ratio for each treatment group. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*
