---
title: "Basic language and operations"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll start off by just introducing some language with basic examples that will 
be used with R. 

## Expression 

An **expression** is a line of code that just evaluates. Nothing is saved - the output 
is simply evaluated and printed. For example, we can perform the operation 5 + 5 and 
print the result in R. 

```{r}
5 + 5
```

## Variable 

We can save **expressions** into **variables**. We can do this using the -> symbol. 
For example, the line of code below saves 5 + 5 into a new variable `my_num`: 

```{r}
my_var <- 5 + 5 

my_var
```

After we save a variable, we can reference and perform operations on this variable 
later on in our code: 

```{r}
my_var + 10
```

## Variable types 

We can save different types of objects into variables. Two of the most common types of
objects are **numeric** variables and **string** variables. Numeric variables are 
numbers; string variables are text. 

```{r}
my_num <- 5 

my_str <- "This is my string"
```

## Vector 

A vector is an ordered collection of numeric or string objects. We can create
a vector by using the `c()` syntax. For example, the code below creates a vector of 
5 numbers, from 1 to 5. 

```{r}
my_vec <- c(1,2,3,4,5)
```

We can select specific elements of the vector by calling the vector, followed by 
`[]`. For example, this line of code selects the first element:  

```{r}
my_vec[1]
```

This line of code selects the fourth element of the vector: 

```{r}
my_vec[4]
```

## Function 

The thing(s) that you put into a function are called argument(s). 

## Dataframe 

## Comments

## Exercises

1. 

2. 

