---
title: "Basic language and operations"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll start off by just introducing some language with basic examples that will 
be used with R. 

## Expression 

An **expression** is a line of code that just evaluates. Nothing is saved - the output 
is simply evaluated and printed. For example, we can perform the operation 5 + 5 and 
print the result in R. 

```{r}
5 + 5
```

## Variable 

We can save **expressions** into **variables**. We can do this using the -> symbol. 
For example, the line of code below saves 5 + 5 into a new variable `my_num`: 

```{r}
my_var <- 5 + 5 

my_var
```

After we save a variable, we can reference and perform operations on this variable 
later on in our code: 

```{r}
my_var + 10
```

Note that doing this operation does not change our variable. If we take a look at 
`my_var`, it still equals 10. 

```{r}
my_var
```

We can change our variable by assigning the result of `my_var + 10` to `my_var`. 

```{r}
my_var <- my_var + 10 

my_var
```


## Variable types 

We can save different types of objects into variables. Two of the most common types of
objects are **numeric** variables and **string** variables. Numeric variables are 
numbers; string variables are text. 

```{r}
my_num <- 5 

my_str <- "This is my string"
```

## Vector 

A **vector** is an ordered collection of numeric or string objects. We can create
a vector by using the `c()` syntax. For example, the code below creates a vector of 
5 numbers, from 1 to 5. 

```{r}
my_vec <- c(1, 2, 3, 4, 5)
```

We can select specific elements of the vector by calling the vector, followed by 
`[]`. For example, this line of code selects the first element:  

```{r}
my_vec[1]
```

This line of code selects the fourth element of the vector: 

```{r}
my_vec[4]
```

Vectors can also consist of strings. 

```{r}
my_str_vec <- c("a", "b", "c")

my_str_vec[1]
```

## Function 

A **function** is an object that does a pre-determined set of operations. A function 
looks something like `function()`, and you put the arguments inside the parenthesis. 

R has a lot of pre-built functions. One example is the function `mean()`. Take a look 
at the below code. The `mean()` function takes in as an argument a vector of numbers, 
and returns the mean of the vector.

```{r}
my_numbers <- c(1, 2, 4, 5, 6)

mean(my_numbers)
```

Another function is `median()`, which will give us the median of a vector of numbers: 

```{r}
median(my_numbers)
```

For all prebuilt functions in R, you can find the documentation online. The simplest way
is by Googling it. [Here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean) 
is the documentation for the `mean()` function.

## Dataframe 

A **dataframe** is a data object in `R`. There are several dataframes included when you download R. 
For example, the `cars` dataset is one that tabulates the speed and distance of different cars' journeys. 

We can use the `head()` function to look at the first few rows of the `cars` dataset.

```{r}
head(cars)
```

We can also read in dataframes. If your data is in a csv format, then you can use 
the `read.csv()` function. It looks something like this:  

```{r}
my_data <- read.csv("crisp-2024-sample100.csv")

head(my_data)
```

## Comments

Code that has a # before it is a **comment** and will not run. Comments are often used 
for documentation - so that you know what code does after you write it. 

```{r}
# This line of code finds the mean of a vector(1,4,6)
mean(c(1, 4, 6))
```

## Exercises

Write code to find the mean of 1, 10, 40, 80, 120, 150. How about the median? 
Standard deviation? Maximum? Minimum? (Hint: look up functions for the standard
deviation, maximum, and minimum.)

