---
title: "Dataset operations and subsetting"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset operations

We start off by reading in the crisp-2024-sample100.csv dataset into our R session. 

```{r}
df <- read.csv("crisp-2024-sample100.csv")
```

After reading in the data, a good first step is to just take a peak at this dataset. 
We can do this using the `head()` function: 

```{r}
head(df)
```

We'll start off by getting to know the dataset. We may want to count the number 
of rows. We can use the `nrow()` function to do this: 

```{r}
nrow(df)
```

Next, we can pull a specific column from the dataset as a vector by using the dollar sign $ operator. For example, we can pull out the ages and save as the `age_vec` variable. 

```{r}
age_vec <- df$age

age_vec
```

Next, we can get some summaries of the column. For example, we can obtain the 
mean age by using the `mean()` function: 

```{r}
mean(age_vec)
```

Note that we combine this into a single step by just directly taking the mean 
of `age` column in `df`: 

```{r}
mean(df$age)
```

What happens if the column of interest has missing data (which are coding as 
NA in R)? Let's take a look, using the `mhdx` column as an example: 

```{r}
mean(df$mhdx)
```

It will give NA as the result! What should we do in this case? We could take the 
mean of the non-missing data. To do this, we can add an extra argument `na.rm = TRUE`:

```{r}
mean(df$mhdx, na.rm = TRUE)
```

Similarly, we can use the `median()`, `sd()`, `min()`, and `max()` functions to 
obtain other descriptive statistics: 

```{r}
# median 
median(df$age)

# standard deviation
sd(df$age)

# min
min(df$age)

# max 
max(df$age)
```

What if we want to obtain multiple summaries of the data at once? We can use
the `summary()` function to obtain the minimum, 25th percentile, median, 
mean, 75th percentile, and maximum of the data, with one line of code: 

```{r}
summary(df$age)
```

What about categorical data? For example, the treatment `tx` variable takes the value 
of 0 or 1, corresponding to BUP-NX and XR-NTX, respectively. How can we get counts 
of each category? We can use the `table()` function to obtain these counts: 

```{r}
table(df$tx)
```

If we use `table()`, it will not tabulate the missing data (NAs). To show the missing data, we have to add an extra argument `useNA = "always"` to our function call: 

```{r}
# tabulating the mhdx column, don't show NAs
table(df$mhdx)

# tabulating the mhdx column, show NAs
table(df$mhdx, useNA = "always")
```

## Subsetting datasets

Subsetting datasets refers to filtering the dataset, by either selecting a set of 
the dataset columns or filtering the rows based on some logic. In our short course, 
we will subset the data using the `dplyr` package. First, we read in the `dplyr` package, which you should have already installed: 

```{r, warning = F}
library(dplyr)
```

The `dplyr` package using a special syntax called **piping**. This involves using a 
special symbol that looks like this `%>%`. This is equivalent to applying a function 
to an object. For example, to find the mean age, we could use the following code: 

```{r}
mean(df$age)
```

Alternatively, we could write the following code, using this special syntax: 

```{r}
df$age %>% 
  mean()
```

This line of code says to take the `df$age` column and apply the `mean()` function to it. Functionally, this is the same as doing `mean(df$age)`. We'll use this syntax when manipulating datasets. 

### Selecting specific columns

Suppose that we want to reduce
the dataset `df` to a specific set of columns - id, tx, age, and sex. We can use 
the `select()` function in the `dplyr` package:

```{r}
# select id, tx, age, and sex
# then, view the first few rows using head()
df %>% 
  select(id, tx, age, sex) %>% 
  head() 
```
Note that this will NOT change the original `df` object. If we look at `df`, we can
still see that it has all of the original columns: 

```{r}
df %>% 
  head()
```

If we want to save the dataframe with the select columns, we can save it as a new 
object: 

```{r}
# create new variable df_select_columns
# which is a dataframe with only the id, tx, age, and sex columns
df_select_columns <- df %>% 
  select(id, tx, age, sex)

# view the first few rows 
df_select_columns %>% 
  head()
```

### Selecting specific rows

Next, we will learn how to select specific rows of a dataframe. We can filter 
using the `filter()` function based on some logic. For example, suppose that we 
want to create a dataframe which only includes BUP-NX (where tx takes the value 
of 0) participants: 

```{r}
df %>% 
  filter(tx == 0) %>% 
  head()
```

Inside the `filter()` function, we specify the logic. In this example, you can see
that we specify `tx == 0`, which means that we only want to include treatment = 0 
participants in our filtered dataset. 

Note that we use a double equal sign `==`, NOT a single equal sign `=` in this logic statement. 

Some other examples: 

1. Greater or less than (< or >): 

```{r}
# filter for age > 30 participants 
df %>% 
  filter(age > 30) %>% 
  head()
```

2. Not (using !)

```{r}
# filter for sex NOT equal to 1  
# use != to say "not equal to 
df %>% 
  filter(sex != 1) %>% 
  head()
```

3. Composite logic using and (&) / or (|)

```{r}
# filter for male and BUP-NX participants 
df %>% 
  filter(sex == 0 & tx == 0) %>% 
  head()
```

```{r}
# filter for male OR BUP-NX participants
df %>% 
  filter(sex == 0 | tx == 0) %>% 
  head()
```
4. Missing rows using (`is.na()`)

```{r}
# select rows that are missing their mhdx value
df %>% 
  filter(is.na(mhdx))
```

```{r}
# select rows that are NOT missing their mhdx value
df %>% 
  filter(!is.na(mhdx)) %>% 
  head()
```


## Exercises 

1. Explain what these lines of code do: 

```{r}
df_new <- df %>% 
  filter(!is.na(mhdx) & age > 30 & tx == 0) %>% 
  select(id, tx, age, mhdx)
```

2. Explain what these lines of code do: 

```{r}
df %>% 
  filter(sex == 0) %>% 
  nrow()
```

3. Explain what these lines of code do: 

```{r}
df_new2 <- df %>% 
  filter(sex == 0)

summary(df_new2$age)

```

4. Explain what these lines of code do: 

```{r}
df %>% 
  filter(is.na(mhdx)) %>% 
  nrow()
```


5. Find the mean BMI of males in the dataset. How about females? 

6. What proportion of those under 30 used marijuana? How about those over 30? If 
there is missing data, then please report the proportion among those not missing 
the relevant data.

6. What proportion of males in the tx = 1 arm have insurance? How about females
in the tx = 1 arm? 


&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*


