---
title: "Basic survival analysis"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Introduction 

In survival data, we not only know if whether someone had the event, we also 
have the **time to the event**. With survival data, we often deal with **censoring**. 
Censoring means that we don't know *when* the event occurred. For some events, 
we only know that it occurred *after* a certain time (this is called **right censoring**). 

With this type of data, we may want to answer the following questions: 

* What is the median survival time? 
* What proportion survived after 1 year? After 2 years? 
* Is the survival different between different groups? 

In this tutorial, we will learn the basics of analyzing survival data in R. This 
will not be comprehensive! The goal is to get a taste of implementing some common 
survival analysis tasks in R. 

## Creating a time to event variable

To analyze survival data in R, we can use the `survival` package. We will also 
use the `dplyr` package in some of the code.

```{r, warning = F, message = F}
# install.packages("survival")

library(survival)
library(dplyr)
```

For this tutorial, we will use the `lung` dataset, which is automatically loaded through 
the survival package. The data contain subjects with advanced lung cancer from the North Central
Cancer Treatment Group. The *event* of interest is death. 

```{r}
head(lung)
```
This dataset has several columns, but we will focus only on time, status, and 
sex for this tutorial: 

```{r}
lung %>% 
  select(time, status, sex) %>% 
  head()
```

Here is the definition of these variables: 

* **Status**: 1-censored (e.g. did not have the event) or 2-dead 
* **Time**: Time to event or censoring 
* **Sex**: 1-male or 2-female 

For those that are censored, we assume that either the study ended or they were no 
longer observed after the censoring date. We know that they had the event (e.g. died)
at some point, but we don't know when. We only know it happened after a specific time. 

To analyze survival data using the `survival` package, we want to make sure there
are two key columns: (1) a time column, and (2) a status column. The status
column must be coded as 0s or 1s, where 0 is for those that are censored and 1 
is for those that were not (e.g. had the event). As a first step, we convert 
the status column in our dataset (which consists of 1s and 2s) to 0s and 1s. 

```{r}
# use the mutate function to change a column of the datset 
lung_new <- lung %>% 
  mutate(status = ifelse(status == 2, 1, 0))
```

Our next step is to create a time-to-event column. This column will encode the information 
in both the time and status columns. We can use the `Surv()` function from the 
`survival` package to create this column. The first argument of this function is the 
time variable, and the second argument is the status variable. 

```{r}
# use mutate function to create time-to-event column
lung_new <- lung_new %>% 
  mutate(tte = Surv(time, status))
```

Let's take a peak at what this new variable looks like: 

```{r}
lung_new %>% 
  select(time, status, tte) %>% 
  head()
```
For those that had the event (status = 1), we can see the tte column shows of 
the time of the event. For those that were censored, we can see that the tte column 
shows the last observation time with a + symbol, indicating that the had the event 
*after* that timepoint. 

## Creating survival plots 

After creating the survival variable, we may want to create a Kaplan-Meier survival
plot. This plot shows the estimated proportion who are still alive at 
different timepoints. To create this plot, we first can use the `survfit()` function. 
The function takes in two arguments The first is the formula which looks like `tte ~ -1`. 
Before the tilde (~), we have tte, which is the time-to-event variable that we 
created. The -1 after the tilde here indicates that we want an overall Kaplan-Meier 
curve, not a stratified one. The second argument is the data. 

```{r}
# first argument is the formula tte ~ -1 
# second argument is the data 
km <- survfit(tte ~ -1, data = lung_new)
```

Next, we can graph this plot using the `ggsurvplot()` function from the `survminer`
package. We put in the survfit object as the argument. 

```{r, warning = F, message = F}
# install.packages("survminer")

library(survminer)
ggsurvplot(km)
```

From this curve, here are some observations: 

* The median survival time (50th percentile) is around 275 days. 
* After 500 days, we estimate the survival probability to be around 30 percent. 

Now suppose that we want to graph the curve, stratified by sex. Within the `survfit()` 
function, we specify the formula `tte ~ sex`. Including sex after the tilde 
creates curves stratified by sex. 

```{r}
# stratified Kaplan-Meier curve 
km_sex <- survfit(tte ~ sex, data = lung_new)

# graph stratified curves 
ggsurvplot(km_sex)
```

This shows the curves separated by males and females. The curves show that females
generally have better survival than males throughout the study. 

## Log rank test

In the previous section, we saw that the females had a noticeably different 
Kaplan-Meier survival curve compared to males. We can perform a hypothesis 
test to check these curves are *significantly* different. The hypothesis test 
to do this is called the **log rank test**. 

The null hypothesis of the log rank test is that the survival curves are the same. 
Setting a significance threshold of 0.05, if we observe a p-value less than that, 
then we reject the null hypothesis (i.e. find evidence of significant differences 
between the two curves).

To perform the log rank test in R, we can use the `survdiff()` function. The 
structure of this function is similar to `survfit()` function. We first 
specify the formula, in this case `tte ~ sex`. This tells the function to test 
if the time-to-event survival variable is significantly different by sex. The
second argument is the data. 

```{r}
survdiff(tte ~ sex, data = lung_new)
```

With $p=0.001$, we find evidence that the two curves are significantly different. 

## Exercise

Using the lung dataset, graph separate Kaplan-Meier curves for participants 
with different values of ph.ecog. (Note: ph.ecog is the ECOG performance score as 
rated by the physician. Ranked 0-3, with increasing severity.) 

Perform a test to see if the survival curves are significantly different.  

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*

