---
title: "Comparing means and proportions"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Comparing means 

One common statistical task is to examine two groups of values and determine if
the means are equal.

Here's an example using the crisp-2024-sample100.csv dataset. Suppose we would like 
to know if the age is different between the two treatment groups. 

We can start by taking a look at what these means are: 

```{r, warning = F, message = F}
# read dplyr library 
library(dplyr)

# read crisp-2024-sample100.csv dataset 
df <- read.csv("crisp-2024-sample100.csv")

# create dataframe with just tx == 0 participants
df_tx0 <- df %>% 
  filter(tx == 0)

# create dataframe with just tx == 1 participants
df_tx1 <- df %>% 
  filter(tx == 1)

# find mean age for tx == 0 participants
mean(df_tx0$age)

# find mean age for tx == 1 participants
mean(df_tx1$age)
```

So the mean age of the `tx == 0` is slightly lower than the `tx == 1` group 
(32.39 versus 33.09). Is this difference significantly different? 

### Hypothesis testing

To determine if this is a statistically significant difference, we can perform a
hypothesis test. For hypothesis testing, we start off by specifying a null hypothesis, 
which is the statement that we want to disprove. So if our question is: "Are the 
mean ages in the two treatment groups different?" then our null hypothesis ($H_0$) would be that the mean ages in the two treatment groups are the same: 

$H_0$: Mean age in `tx == 0` group EQUALS mean age in the `tx == 1` group

The alternative hypothesis ($H_a$) is the opposite, which is: 

$H_a$: Mean age in tx == 0 group does NOT equal the mean age in the tx == 1 group

There are several hypothesis tests to test this hypothesis. One very popular one is called the t-test. To implement the t-test in R, we can use the function `t.test()`. Within `t.test()`, we 
include two arguments for the two vectors of numeric data that we want to test.

```{r}
# perform t-test, which tests the null hypothesis that the
# mean ages are the same between the tx = 0 and tx = 1 groups
t.test(df_tx0$age, df_tx1$age)
```
The output of `t.test` gives us a lot of information. We'll focus on a few aspects of the output: 

* **p-value**: This is the p-value for the hypothesis test. The p-value is the probability of obtaining a test statistic at least as extreme as what we get with our data, given that the null hypothesis is true. So with a p-value of 0.702, we have approximately a 70% probability of obtaining a test statistic at least as extreme as what we get in our data, given the null hypothesis is true. Typically, we set the p-value threshold (or alpha) to 0.05 in our hypothesis test. This means we reject the null hypothesis if our p-value is less than 0.05. Intuitively, we want to reject the null hypothesis if our data is extreme compared to what our null hypothesis says.
* **95 percent confidence interval**: This is a confidence interval for the difference in mean age between the `tx == 0` and `tx == 1` groups. So we estimate that a 95% CI for the difference in age between the `tx == 0` and `tx == 1` groups is (-4.31, 2.91).

In this output, we fail the reject the null hypothesis ($p = 0.702$), meaning that we fail to find evidence that the two groups have a significantly different mean age. 

There are several other hypothesis tests comparing the mean or median of two samples: paired t-test (for paired data), Wilcoxon rank-sum test (to test for equal medians), etc.

## Comparing proportions

Another common statistical task is to examine two proportions and determine if they are equal. 

For example, suppose that we want to know the proportion of people with insurance is the same between the treatment arms. 

The first step to doing this might be to calculate the raw numbers of those with insurance in each treatment arm. 

We can do this by hand using the code below:

```{r}
# number of tx = 0, no insurance
df %>% 
  filter(tx == 0 & insure == 0) %>% 
  nrow()

# number of tx = 0, has insurance
df %>% 
  filter(tx == 0 & insure == 1) %>% 
  nrow()

# number of tx = 1, no insurance
df %>% 
  filter(tx == 1 & insure == 0) %>% 
  nrow()

# number of tx = 1, has insurance
df %>% 
  filter(tx == 1 & insure == 1) %>% 
  nrow()

```

Alternatively, we can create a 2x2 table using the the 
`table()` function, where we include as an input a two-column
dataframe with the columns of interest. 

```{r}
# create 2x2 table from two column, binary variable dataframe
tx_insure_tab <- table(df %>% select(tx, insure))

tx_insure_tab
```

We observe that 41/66 (73%) of those in the `tx == 0` arm have insurance, compared to 36/44 (82%) of those in the `tx == 1` arm. 

```{r}
# proportion in tx == 0 arm 
41/(41 + 15)

# proportion in tx == 1 arm
36/(36 + 8)
```

### Hypothesis testing

The next question to ask is are these proportions significantly different. 
We will use a hypothesis test to test this. The null hypothesis is that the 
proportions are the same. The alternative hypothesis is that the proportions 
are different. 

One option is Fisher's exact time. One option is the two-sample z-test for 
proportions. We can implement this in `R` using `prop.test()`. 

```{r}
prop.test(tx_insure_tab)
```
Another option (better for smaller samples) is the Fisher exact test. We can 
implement this in R using `fisher.test()`: 

```{r}
fisher.test(tx_insure_tab)
```
A third option is the chi-squared test. We can implement this in R using 
`chisq.test()`: 

```{r}
chisq.test(tx_insure_tab)
```
Note that the chi-squared test gives the same p-value result (p = 0.438)  as the 
two-sample z-test for proportion. This is because the two tests are the same! 
The only difference is that the chi-squared test can be used for bigger tables 
(e.g. 3x2, 3x3 tables, etc.). 

In all three tests, we see a p-value greater than 0.05, so we fail to find 
evidence that there is a significant difference in the proportion with insurance 
in each treatment arm.

## Exercises 

1. What is the mean bmi for each sex (males and females)? Perform a hypothesis 
test to see whether the mean bmi is significantly different between sexes.

2. What proportion in each treatment arm have a history of psychotic diagnosis 
or episodes? Are the proportions significantly different? 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*


