---
title: "Using the table1 package"
output:
  html_document:
    toc: true
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## The hard way

For assignment 3, you were asked to create a "Table 1" from a dataset. You may
have done this by hand. It may have looked something like this:

```{r, message = F, warning = F}
# read in dplyr package
library(dplyr)

# read in dataset 
df <- read.csv("crisp-2024-sample100.csv")

# split into BUP-NX and XR-NTX datasets
df_bup <- df %>% filter(tx == 0)
df_xr <- df %>% filter(tx == 1)

# mean age for BUP-NX group 
mean(df_bup$age)

# sd of age for BUP-NX group 
sd(df_bup$age)

# mean age for XR-NTX group
mean(df_xr$age)

# sd of age for XR-NTX group
sd(df_xr$age)

# etc... 
```

The problem with this is that this is a very manual, time-intensive process. 
Alternatively, we can use the `table1` package to help us do this. 

## The easier way

First, let's install the `table1` package if we haven't yet done so. 

```{r, eval = F}
install.packages("table1")
```
Now let's read in the package: 

```{r, message = F}
library(table1)
```
Take some time and read the [documentation](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html) for the `table1` package. This package allows us to summarize multiple variables and output
the summary in a publication-ready table. 

Before we use this package, we will convert our categorical variables 
into factors. We'll start off by doing this for the variables `tx` and `sex`:

```{r}
# convert categorical columns into factors
df$tx <- factor(df$tx, 
                labels = c("BUP-NX", "XR-NTX"))

df$sex <- factor(df$sex, 
                 labels = c("Male", "Female"))

```

This is a necessary first step, or else the `table1` package will this that 
these columns are numeric!

After this, we can run the `table1()` function: 

```{r}
table1(~ age | tx, data = df)
```

The `table1()` function includes two arguments. First, we specify the formula for
how we want the table to look. This formula takes the form `~ var1 | var2`, where
`var1` are the columns that we want to summarize, followed by a `|`, and then `var2`, 
which is the column that we want to stratify by. So `~ age | tx` means that we want to 
summarize the age column, stratified across the two treatment arms. Next, we specify the dataframe, 
which is why we include a `data = df` additional argument. 

We can add variables to summarize by using `+`. The following line of code summarizes 
both the age and sex variable across treatment arms: 

```{r}
table1(~ age + sex | tx, data = df)
```


Note that by default, `table1` will provide the mean, median, min, max, and standard 
deviation for continuous numeric variables and proportions for categorical factor variables.

## Exercise

Show descriptive statistics for age, number of years of education, and smoking status 
stratified by sex using the `table1` package. 

&nbsp; &nbsp; &nbsp;

*[Click here](https://jpspeng.github.io/crisp_notes/) to go back to the CRISP R notes homepage.*
