---
title: 'CRISP Day 1 2025: Getting starting with R'
output:
  html_document:
    df_print: paged
---

In today's lesson, we will get our computer set up with R. 

# Installing R 

If you have not yet, please start by downloading and installing R and R Studio. 

1. Download R ([link](https://cran.r-project.org/)) and run the installation file. 
2. Download R Studio ([link](https://posit.co/download/rstudio-desktop/)) and run the installation file. 

# Folder setup 

1. Please create a folder somewhere on your computer called **crisp_R_notes**
(or something similar). 
2. Within this folder, please download and save the data CSV file from the 
[course website](https://jpspeng.github.io/crisp_notes/).
3. Within this folder, please download and save the R Markdown notebook (.Rmd) for today 
(July 7). For future lessons, please put all notebooks within this folder. 

# Creating, saving, and opening R files  

1. Open R Studio. Click on File -> New File -> R Script. Then, press File -> Save As, 
and save the file as *test.R* in your crisp_R_notes folder. 
2. Click on File -> New File -> R Notebook. Then, press File -> Save As, 
and save the file as *test.Rmd* in your crisp_R_notes folder. 
3. Open today's R Markdown file (*crisp_2025_day1.Rmd*) and continue this 
tutorial within R Studio. 


# Code chunks and comments 

Code within R notebooks run within code chunks. To add a code chunk, you can press
the +C icon on the right hand corner of the Source window, and click on R.

To run the code chunk, **press the play button** on the upper right hand corner of the
code chunk: 

```{r}
5 + 5
```

Code that has a # before it is a comment and will not run. Comments are often 
used for documentation - so that you remember what code does after you write it. 

Try running this line of code, and see what happens: 

```{r}
# 5 + 5 
```


# Installing and using packages 

One of the main advantages of R are its packages. 

An R package is a collection of functions, data, and documentation that extends 
the capabilities of R. They are used to share code and data with others and to 
reuse code in different projects. They can include:

1. **Functions**: Custom functions that perform specific tasks.
2. **Data**: Datasets that can be used for analysis and examples.
3. **Documentation**: Help files and vignettes that explain how to use the functions 
and datasets.

R packages are typically installed from CRAN (Comprehensive R Archive Network). Packages
tend to have specific uses. For example, there are packages specifically for 
making graphs, others for creating manuscript-ready tables, and others to simplify 
data processing. 

You can install a package in R using the `install.packages()` function, and load 
it using `library()`.

```{r}
# Install a package from CRAN
install.packages("dplyr")
```

```{r}
# Load the package
library(dplyr)
```

# Reading in data into R 

Say that you have a dataset. The first step is to read in data into R. You can do
this using the `read.csv()` function. 

The easiest way to use this function is to put the data file in the **same folder** 
as your R (or Rmd) code. You should have done this already with the folder setup 
above. Try to run the following code: 

```{r}
read.csv("crisp-2024-sample100.csv") 
```

Did it run? Then great! That means that the data is now in our R session. We can 
now manipulate and analyze the data, which we will do in the next few tutorials.

If the data is not in the same folder as your R file, you may have to include the
directory path to the data in the argument when calling `read.csv()`.

```{r}
# if data is in a different folder, it may look like this: 
# df <- read.csv("C:/path/to/directory/crisp-2024-sample100.csv")
```

# Exercises 

1. Install the following packages: `ggplot2`, `table1`, and `survival`. 

```{r}
# type in code here
install.package("ggplot2")
install.package("table1")
install.package("survival")

```

2. Use the `+`, `-`, `*`, and `/` operators to:
* Add 1048 and 890
* Multiply the previous result by 20
* Subtract the previous result by 50 
* Divide the previous result by 2.5

```{r}
# type in code here 
1048 + 890
1938 * 20 
38760 - 50
38710 / 2.5
```

3. Describe what the following code is doing: 

**Answer**: The code is doing the same series of operations as Q2.  

```{r}
a = 1048 
b = a + 890
c = b * 20 
d = c - 50 
e = d / 2.5 

print(e)
```

4. Following the pattern in the previous code block, perform the following operations: 
* Multiply 13 and 631
* Divide the previous result by 25 
* Add 10 to the previous result 

```{r}
# type in code here 
a = 13
b = a * 631 
c = b / 25 
d = c + 10 

d
```

5. Describe what the code below is doing: 

**Answer**: The code is adding 5 to each element of the list 1,2,3,4,5. 

```{r}
a = c(1,2,3,4,5)
b = a + 5

print(b)
```

5. Describe what the code below is doing: 

**Answer**: The code is adding 5 to each element of sequence 5,6,7,8,9,10. 

```{r}
a = 5:10
b = a + 5

print(b)
```

6. Use what you learned from the previous questions to multiply all numbers 
from 25 through 31 by 13. 

```{r}
# type in code here 
a = 25:31 
b = a * 13 

print(b)
```





