---
title: 'CRISP Day 6 2026: Linear regression'
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r, warning = F, message = F}
library(dplyr)
```

In today's lesson, we will learn how to run and interpret linear regression.  

```{r}
# read data 
my_data <- read.csv("crisp-2024-sample100.csv")
crisp_survey_data <- read.csv("crisp_survey_data_v2.csv")

# create factor variables 
my_data <- my_data %>% 
  mutate(sex = factor(sex, labels = c("male", "female")))

crisp_survey_data <- crisp_survey_data %>% 
  mutate(sex = factor(sex, labels = c("female", "male")), 
         age_group = factor(age_group, labels = c("under33", "over33")),
         pet_preference = factor(pet_preference, labels = c("dogs", "cats", "both", "neither")), 
         morning_or_evening_person = factor(morning_or_evening_person, labels = c("morning", "evening", "neither")))
```

# Linear regression with continuous predictor 

The goal of linear regression is to relate an outcome to one (or more) predictors. 
For example, **how does number of years of education associate with age, on average?**

The first step in running a linear regression is to write out the model that we
are trying to fit. In this example, here is our working model:

$$
YrsEduc = \beta_0 + \beta_1age 
$$

Where $YrsEduc$ is the participant's number of years of education, 
$age$ is the participant's age, $\beta_0$ is the intercept coefficient, and $\beta_1$ is
the coefficient of interest. 

With linear regression, our goal is to to find the values of $\beta_0$ 
and $\beta_1$ which best fit the data. Behind the scenes, the software will find the 
values of $\beta_0$ and $\beta_1$ which minimize the the squared error between the 
fitted line's values and the true data's values. 

Running this in R is relatively straightforward. We can use the `lm()` function, which 
is a pre-built function in R (no packages needed). The `lm()` function takes in two 
arguments: (1) the formula, which looks like `outcome ~ predictor`, and (2) the data. 
In the code below, we run a linear regression with `educyrs` as our outcome and `age` 
as our predictor, and then save it in the variable `fitted_mod`. 

```{r}
# to run linear regression in R, use lm() 

# first argument in the lm() function is your formula 
# formula looks like outcome ~ predictor 
# outcome: educyrs
# exposure: age 

# second argument is the dataframe  
fitted_mod <- lm(educyrs ~ age,  data = my_data)

# to summarize the fitted model, you can use the summary() function 
summary(fitted_mod)

# to get confidence intervals, use the confint() function 
confint(fitted_mod)
```

```{r}
fitted_mod <- lm(educyrs ~ age, data = my_data)
```

To see the fitted coefficient of this model, we can apply the `summary()` function
on `fitted_mod`. 

```{r}
summary(fitted_mod)
```

We focus in on the Coefficients section of the summary. 

1. The Estimate column tells us the fitted values for the intercept ($\beta_0$) and the 
coefficient on age ($\beta_1$). We estimate the intercept of the best fit line to be
14.30 and the coefficient on age to be -0.064.

2. The Std. Error gives us the standard error on the estimate. Essentially, this is a measure
of how much uncertainty we have in our estimate. Larger standard errors indicate that 
we have less certainty in the estimate. 

3. The last column Pr(>|t|) gives us a p-value for a hypothesis test. Specifically, 
it tests the hypothesis that the coefficient is equal to 0. For example, the p-value 
on the age coefficient is 0.00573, so if we use 0.05 as our testing threshold, then we
reject the null hypothesis that the coefficient on age is 0. 

Next, we can obtain confidence intervals using the `confint()` function: 

```{r}
confint(fitted_mod)
```

This will give us 95% confidence intervals on the coefficients. For example, 
this tells us that a 95% confidence interval for the coefficient on age is 
(-0.109, -0.019).

How can we interpret this in plain language? 

*We estimate that for two groups differing in age by one year, the older group has 
on average 0.064 years fewer education (95% CI: 0.019, 0.109). At the 5% level,
we reject the null hypothesis that the coefficient on age is 0. That is, we find 
evidence of an association between age and years of education.* 

**Example**: In the CRISP survey data, regress anxiety/depression/irritability score (`anxious_depressed_or_irritable`) on hours of sleep (`hours_of_sleep`). Interpret your result. 

```{r}
crisp_mod <- lm(anxious_depressed_or_irritable ~ hours_of_sleep, data = crisp_survey_data)

summary(crisp_mod)
confint(crisp_mod)
```

*Interpretation*: We estimate that someone with 0 hours of sleep will have a 
anxiety score of 4.5 (95% CI: 2.00, 7.01). 

Comparing two groups differing in one hour of sleep, we estimate that the group 
with more sleep will have an anxiety score of 0.25 less (95% CI: -0.61, 0.11)
than the group with less sleep. We do not find a significant association between 
hours of sleep and anxiety score (p = 0.17).

# Linear regression with binary predictor 

Suppose that we are examining number of years of education (`educyrs`) as our outcome, 
and binarized sex (`sex`) as our single predictor. 

Note that `sex` takes the value of 0 or 1 in our dataset, with 0 representing males
and 1 representing females. 

We aim to find the best linear fit considering these data points. That is, we aim 
to fit the model: 

$$
educyrs = \beta_0 + \beta_1 sex
$$

We perform a linear regression, regressing `educyrs` on `sex` using the `lm()`
function. 

```{r}
fitted_mod2 <- lm(educyrs ~ sex, data = my_data)

summary(fitted_mod2)
```

How can we interpret this output?

* *Intercept*: The mean number of years of education for males (or those with 
sex = 0). That is, males have on average 12.21 years of education. 
* *Coefficient on sex*: The difference in the mean number of years of education 
between females and males. That is, we estimate that females have on average 0.039 
fewer years of education than males. 
  * *p-value*: The result from testing the null hypothesis that the coefficient on 
  sex is equal to 0. Equivalently, we are testing the null hypothesis that there is 
  no difference in number of years of education between males and females. With
  $p = 0.93$, we fail to find evidence of a difference in the number of years of 
  education comparing males and females. 

We can also obtain confidence intervals on these estimates using `confint()`. 

```{r}
confint(fitted_mod2)
```

**Example**: In the CRISP survey data, regress weekly minutes of cardio exercise
(`cardio_exercise`) on age group (`age_group`). Interpret your result. 

```{r}
# write code here 
crisp_mod2 <- lm(cardio_exercise ~ age_group, data = crisp_survey_data)

summary(crisp_mod2)
confint(crisp_mod2)
```
*Interpretation*: We estimate that those over age 33 exercise 98 minutes 
(95% CI: 25, 171) less on average per week compared to those under age 33. 
We reject the null that the mean cardio exercise time is the same between the 
two groups (p = 0.01).

# Linear regression with categorical variable 

Suppose that we have a categorical predictor. If we run linear regression, then 
the output displays the comparison between each group and the reference group 
(which defaulted as the lowest factor group).

**Example**: In the CRISP survey data, regress weekly minutes of cardio exercise
(`cardio_exercise`) on pet preference (`pet_preference`). Interpret the coefficient on 
the cats coefficient. 

```{r}
crisp_mod3 <- lm(cardio_exercise ~ pet_preference, data = crisp_survey_data)

summary(crisp_mod3)
```
*Interpretation (of the cats coefficient*): We estimate that those who prefer 
cats exercise 59 minutes more on average per week compared to those that prefer dogs. 

Note: To specify the order of factor variables, you can use the `levels` argument
when you are creating the factor variable. For example: 

```{r}
crisp_survey_data <- crisp_survey_data %>% 
  mutate(pet_preference = factor(pet_preference, levels = c("cats", "dogs", 
                                                            "both", "neither")))

crisp_mod4 <- lm(cardio_exercise ~ pet_preference, data = crisp_survey_data)

summary(crisp_mod4)
```

# Linear regression with more than one predictor

Suppose that we run a linear regression, with number of years of education `educyrs` 
as the outcome, and `age` and `sex` as the predictors. That is, we are trying 
find the best estimates of $\beta_0$, $\beta_1$, and $\beta_2$ in the equation below: 

$$
educyrs = \beta_0 + \beta_1 age + \beta_2 sex
$$

To do this in R, we can use `lm()` with `age + sex` as our predictors in the formula: 

```{r}
fitted_mod3 <- lm(educyrs ~ age + sex, data = my_data)

summary(fitted_mod3)

confint(fitted_mod3)
```

How can we interpret the output? 

* *Intercept*: The predicted number of years of education for males (sex = 0) with
age = 0. Note that this is not meaningful, since we don't have age = 0 in our
population of interest. 
* *Coefficient on age*: The estimated difference in number of years of education comparing 
two groups *of the same sex* differing by one year of age. 
* *Coefficient on sex*: The estimated difference in the number of years of education 
comparing females versus males *of the same age*. 

*Example*: In the CRISP survey data, regress anxiety/depression/irritability score (`anxious_depressed_or_irritable`)
on hours of sleep (`hours_of_sleep`) adjusting for sex (`sex`). Interpret your result. 

```{r}
# write code here 

```

# Exercises 

For the exercises below, use the sample dataset (`my_data`). 

1. We are interested in learning about the association between BMI (`bmi`) and 
smoking status (`smoke100`). Run a linear regression with `bmi` as the outcome 
and `smoke100` as the regressor. Provide an interpretation of the results. 

```{r}
# write code here 

```

2. We believe that the relationship between BMI and smoking status is confounded 
by age. Adjust for age in your linear regression in question 1. Run the regression 
and provide an interpretation of the results. 

```{r}
# write code here 

```

3. We wish to run the same regression from question 2 separately among men and 
women. Perform two separate regression below for both sexes. Provide an interpretation
for your results. 

```{r}
# write code here 

```

