---
title: 'CRISP 2026: Homework 1'
output: html_document
---

1. Load in the library `dplyr` into R. 

```{r}
# write code here 

```

2. Download in the `hw1_fake_data.csv` dataset, load it into R, and save it as 
the variable `df`. 

```{r}
# write code here 

```

Here is the data dictionary for this dataset:

* `id`: Unique person identifier.
* `treatment`: Treatment group indicator; `0` = control, `1` = treatment.
* `gender`: Participant gender; `male` or `female`.
* `weight`: Participant weight in pounds. Some values are missing.
* `diabetes`: Diabetes status indicator; `0` = no diabetes, `1` = has diabetes.

3. Count the number of rows in the dataset. 

```{r}
# write code here 

```

4. How many females are in the dataset? How many participants are on treatment? 
How many participants have diabetes? 

```{r}
# write code here 

```

5. Provide summary statistics for the weights of all participants. How many 
participants are missing the `weight` variable? 

```{r}
# write code here 

```

6. Subset the dataset into participants who have diabetes and participants who 
don't have diabetes, selecting only the columns `id` and `diabetes`. Assign these 
into variables with names `diabetes_df` and and `no_diabetes_df`, respectively. 

```{r}
# write code here 

```

7. Give summary statistics for participants' weights comparing those with 
diabetes versus those without. 

```{r}
# write code here 

```

8. Comparing treated versus untreated participants: 
* How many participants are in each group? 
* How many participants have diabetes in each group? 
* How many participants are male in each group? 
* What proportion in each group have diabetes? 
* What proportion in each group are male? 
(Hint: Use `group_by()`, `summarise()`, and `mutate()`)

```{r}
# write code here 

```
