---
title: 'CRISP Day 5 2026: Creating Table 1s and comparing means/proportions'
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
---

```{r, warning = F, message = F}
library(dplyr)
```

```{r}
my_data <- read.csv("crisp-2024-sample100.csv")
```

## Introduction 

Last time, we learned how to construct summaries of subgroups of our data. To do
this, we first `group_by()` the subgroup of interest and then define our summary 
in the `summarise()` function. For example: 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(mean_age = mean(age), 
            sd_age = sd(age), 
            mean_bmi = mean(bmi, na.rm = T), 
            sd_bmi = sd(bmi, na.rm = T),
            num_males = sum(sex == 0))
```

What if we want to do this for several variables? Doing this manually 
will take a long time! Today, we are going to learn a shortcut for doing this 
using the `table1` package. 

## Factor variables 

Before we get into learning how to use the `table1` package, we will need to 
first learn about **factor** variable types. 

Each dataframe column can have different types. Some variable are *string* variables 
(e.g. text); others are *numeric* variables (e.g. numbers). There is another type 
of variable in R, called a **factor**. These are categorical variables, which often 
are coded as integers in a dataset (e.g. 0, 1, 2, etc.). 

One example: we may want to consider the variable `tx` in the crisp-2024-sample100.csv 
dataset as a factor variable because it only takes two distinct values: 0 and 1, where
0 corresponds to BUP-NX and 1 corresponds to XR-NTX. 

```{r}
table(my_data$tx)
```

By default, R will read in number variables as in *numeric*. It starts off by 
thinking this column is just a number. But, we may want to change it into a 
factor before we perform our analysis. To do this, we can use the following line of code: 

```{r}
my_data %>% 
  mutate(tx_factor = factor(tx, labels = c("BUP-NX", "XR-NTX"))) %>% 
  select(tx, tx_factor)
```

**Example:** Complete the following code to convert the following variables 
into factor variables with the following labels: 

* `sex`: 0 = males, 1 = females 
* `insure`: 0 = uninsured, 1 = insured
* `smoke100`: 0 = no, 1  = yes
* `mhdx`: 0 = no, 1 = yes

```{r}
# complete the "FILL IN HERE" blanks in the code below 
my_data_factor <- my_data %>% 
  mutate(tx = factor(tx, labels = c("BUP-NX", "XR-NTX")),
         insure = factor(insure, labels = c("uninsured", "insured")),
         sex = "FILL IN HERE", 
         smoke100 = "FILL IN HERE", 
         mhdx = "FILL IN HERE") 

```

**Note**: Remember the `summary()` function? Once we convert the variables into 
factors, we can now use it to correctly summarize categorical variables in the data. 

```{r}
summary(my_data_factor)
```

## Using `table1` package

Let's read in the package: 

```{r, message = F}
library(table1)
```

Before we use this package, we will need to convert our categorical variables 
into factors. Luckily, we already did this in the previous example. This is a necessary 
first step if there are coded categorical variables, or else the `table1` 
package will this that these columns are numeric!

After this, we can run the `table1()` function: 

```{r}
table1(~ age | tx, data = my_data_factor)
```

The `table1()` function includes two argument: 

* First, we specify the formula for
how we want the table to look. This formula takes the form `~ var1 | var2`, where
`var1` are the columns that we want to summarize, followed by a `|`, and then `var2`, 
which is the column that we want to stratify by. So `~ age | tx` means that we want to 
summarize the age column, stratified across the two treatment arms. 
* Next, we specify the dataframe, which is why we include a `data = my_data_factor` additional argument. 

We can add variables to summarize by using `+`. The following line of code summarizes 
both the age and sex variable across treatment arms: 

```{r}
table1(~ age + sex | tx, data = my_data_factor)
```

Note that by default, `table1` will provide the mean, median, min, max, and standard 
deviation for continuous numeric variables and proportions for categorical factor variables.

We can also obtain the overall summary statistics by excluding the `|` notation. 
For example: 

```{r}
table1(~ age + sex, data = my_data_factor)
```

We can change the labels in these Table 1's by setting labels 
for each column of interest. To do this, the code looks like this: 

```{r}
label(my_data_factor$sex) <- "Sex"
label(my_data_factor$age) <- "Age"

table1(~ age + sex | tx, data = my_data_factor)

```

**Example**: Write code to create a table summarizing the age (`age`), 
number of years of education (`educyrs`), insurance statue (`insure`), 
and smoking status (`smoke100`) stratified by treatment `tx` 
using the `table1` package. 

```{r}
# write code here
```

## Comparing means 

Table 1's often include p-values which tell us whether there is a significant 
difference between groups using a hypothesis test. 

For hypothesis testing, we start off by specifying a null hypothesis, 
which is the statement that we want to disprove. So if our question is: "Are the 
mean ages in the two treatment groups different?" then our null hypothesis ($H_0$) 
would be that the mean ages in the two treatment groups are the same: 

$H_0$: Mean age in `tx == 0` group EQUALS mean age in the `tx == 1` group

The alternative hypothesis ($H_a$) is the opposite, which is: 

$H_a$: Mean age in tx == 0 group does NOT equal the mean age in the tx == 1 group

There are several hypothesis tests to test this hypothesis. One very popular one is 
called the t-test. To implement the t-test in R, we can use the function `t.test()`. 
Within `t.test()`, we specify a formula which looks like `variable ~ group`. 
In this case, we want to compare the the `age` across `tx`, so we specify 
`age ~ tx`. In the second argument, we specify the dataframe to reference. 

```{r}
# perform t-test, which tests the null hypothesis that the
# mean ages are the same between the tx = 0 and tx = 1 groups
t.test(age ~ tx, data = my_data_factor)
```

* **p-value**: This is the p-value for the hypothesis test. The p-value is the 
probability of obtaining a test statistic at least as extreme as what we get with 
our data, given that the null hypothesis is true. So with a p-value of 0.702, 
we have approximately a 70% probability of obtaining a test statistic at least 
as extreme as what we get in our data, given the null hypothesis is true. 
Typically, we set the p-value threshold (or alpha) to 0.05 in our hypothesis 
test. This means we reject the null hypothesis if our p-value is less than 0.05.
Intuitively, we want to reject the null hypothesis if our data is extreme compared 
to what our null hypothesis says.
* **95 percent confidence interval**: This is a confidence interval for the difference 
in mean age between the `tx == 0` and `tx == 1` groups. So we estimate that a 
95% CI for the difference in age between the `tx == 0` and `tx == 1` groups is 
(-4.31, 2.91).

There are several other hypothesis tests comparing the mean or median of two 
samples: paired t-test (for paired data), Wilcoxon rank-sum test (to test for 
equal medians), etc.

**Example:** Write code to perform a hypothesis test, testing if there is a 
difference between the mean number of years of education (`educyrs`) between
the two treatment (`tx`) arms. 

```{r}
# write code below 
```

## Comparing proportions 

Another common statistical task is to examine two proportions and test 
if they are equal. 

For example, suppose that we want to know the proportion of people with 
insurance is the same between the treatment arms. 

The first step to doing this is to create a 2x2 table using the the 
`table()` function, where we include as an input a two-column dataframe with the 
columns of interest.

```{r}
# create 2x2 table from two column, binary variable dataframe
tx_insure_tab <- table(my_data_factor %>% select(tx, insure))

tx_insure_tab
```

Then, we will use a hypothesis test to test this. The null hypothesis is that the 
proportions are the same. The alternative hypothesis is that the proportions 
are different. 

One option is the two-sample z-test for proportions. We can implement this in 
`R` using `prop.test()`. 

```{r}
prop.test(tx_insure_tab)
```

Another options is the chi-squared test. We can implement this in R using 
`chisq.test()`: 

```{r}
chisq.test(tx_insure_tab)
```

Note that the chi-squared test gives the same p-value result (p = 0.438)  as the 
two-sample z-test for proportion. This is because the two tests are the same! 
The only difference is that the chi-squared test can be used for bigger tables 
(e.g. 3x2, 3x3 tables, etc.). 

**Example:** Write code to perform a hypothesis test, testing if there is a 
difference between the proportion of smokers and non-smokers (`smoke100`) between
the two treatment (`tx`) arms. 

```{r}
# write code below 
```

## Exercises

Let's explore the CRISP survey data using the tools that you have learned so far! 

* `record_id`: Unique REDCap survey record number.
* `age_group`: Age category (`0` = 33 or younger; `1` = older than 33).
* `sex`: Sex (`1` = Woman; `2` = Man; `3` = Non-binary; `4` = Prefer not to say).
* `pet_preference`: Pet preference (`1` = Dogs; `2` = Cats; `3` = Both; `4` = Neither).
* `morning_or_evening_person`: Time-of-day preference (`1` = Morning; `2` = Evening; `3` = Equal or neither).
* `food_preference`: Food preference (`1` = Gourmet; `2` = Upscale; `3` = Home cooking; `4` = Fast food).
* `spent_time_in_nature`: Spent at least two hours in nature during the past week (`0` = No; `1` = Yes).
* `interacted_with_clubs`: Interacted with at least one club or organization during the past week (`0` = No; `1` = Yes).
* `resistance_training_workouts`: Number of resistance-training workouts during the past week (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `hours_of_sleep`: Average hours of sleep per night (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `packaged_snacks_per_day`: Average number of packaged snacks consumed per day (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `anxious_depressed_or_irritable`: Frequency of feeling anxious, depressed, or irritable (`1` = Never; `2` = Rarely; `3` = Sometimes; `4` = Often; `5` = Always).
* `fatigue_rating`: Average fatigue rating (`1` = None; `2` = Mild; `3` = Moderate; `4` = Severe; `5` = Very severe).

```{r}
crisp_survey_df <- read.csv("crisp_survey_data_v1.csv")
```

1. Convert the `spent_time_in_nature` and `food_preference` variables into 
factor variables. 

```{r}
# write code here 

```

2. Create a Table 1 examining the variables `hours_of_sleep`, 
`anxious_depressed_or_irritable`, and `food_preference`, stratifying by people who 
spent time in nature versus those who haven't. 

```{r}
# write code here 

```

3. Perform a hypothesis test comparing the hours of sleep for those who spent
time in nature versus those who haven't. What is the null hypothesis for this 
hypothesis test? Interpret the result. 

```{r}
# write code here 

```

4. Perform a hypothesis test comparing the anxiety/depression score for those who spent
time in nature versus those who haven't. What is the null hypothesis for this 
hypothesis test? Interpret the result. 

```{r}
# write code here 

```

5. Is there evidence that the proportion of men who spend time in nature differ
from the proportion of women that spend time in nature in this dataset? 

```{r}
# write code here 

```


