---
title: 'CRISP Day 3 2025: Dataframe processing and subsetting'
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
library(dplyr)
```

In today's lesson, we will learn about how to do basic dataset processing
using both base R and the `dplyr` package. 

```{r}
my_data <- read.csv("crisp-2024-sample100.csv")
```

# Changing / creating new columns 

Suppose that we wanted to add new columns or change existing columns of our 
dataset. We can do this by assigning a vector to a new or existing column as
follows: 

```{r}
# create vector of numbers 1 to 100
new_ids_vec <- 1:100

# add new column called new_id which equals the new_ids_vec vector
my_data$new_id <- new_ids_vec

# we can also do this in one step 
my_data$new_id <- 1:100 

# this line of code would replace the existing id column with 1:100 
my_data$id <- 1:100
```

**Example**: Modify the `my_data` dataframe to create a new column called
`age_months` which equals the `age` column multiplied by 12. 

```{r}
# write code below

```

# Subsetting dataframes using base R (optional)

Remember that we can subset (i.e. take select elements from) vector objects using
the bracket `[]` notation. 

```{r}
my_vec <- c(1,3,5,7,9)

# select the second and third elements 
my_vec[c(2,3)]
```

We can also subset dataframes, which means that we take a selection of the columns 
and/or rows. Like subsetting vectors, subsetting dataframes also follows the 
bracket `[]` notation. However, dataframes have two dimensions, so we need to 
specify both row and column subset "instructions". The subset "instructions" are 
separated by a comma: 

* The text before the comma specifies the rows selected
* The text after the comma specifies the columns selected

```{r}
# pull out the id and age of participant in the second row 
my_data[2, c("id", "age")]
```

```{r}
# pull out the id and tx column of participants in the first 10 rows 
my_data[1:10, c("id", "tx")]
```

```{r}
# select the id and tx column for all participants
my_data[, c("id", "tx")]
```

```{r}
# select all columns for the first 10 rows 
my_data[1:10, ]
```

We can also use this functionality to subset rows based on logic. For this, 
we can use the `>, >=, <, <=, ==, !=` (not equal to) logic operators. For example, the 
code below shows how to filter for subjects who are not treated (`tx == 0`) and 
those who are over the age of 30 (`age > 30`).

```{r}
# Select all columns for all rows where tx == 0. 
# NOTE: This is NOT written as my_data[tx == 0,]
# ALSO: It is written as == (double equal) not = 
my_data[my_data$tx == 0,]
```

```{r}
# select the id, educyrs column for those who age > 30 
my_data[my_data$age > 30, c("id", "educyrs")]
```

We can also subset for those who are missing or not missing a specific variable. 
To do this, we can use the `is.na()` logic check command. For example, the code
below filters for all of those who are missing the BMI variable. 

```{r}
my_data[is.na(my_data$bmi), ]
```
Also, the `!` operator means "not" in R. For example, the lines of code below 
count the number of people are **not** missing the `mhdx` variable. 

```{r}
has_mhdx_df <- my_data[!is.na(my_data$mhdx), ]

nrow(has_mhdx_df)
```

**Example**: Subset `my_data` to select only those who are male (sex equals 0), 
and then save this as a new dataframe called `male_data`. Find the mean BMI of 
those who are male (and who are not missing BMI data). 

```{r}
# write code below 

```


# Dataframe manipulation using `dplyr` (my preference!) 

We can also subset data using the `dplyr` package. In my opinion, this is easier 
and more readable than using base R. First, we read in the `dplyr` package, 
which you should have already installed: 

```{r, warning = F}
library(dplyr)
```

The `dplyr` package using a special syntax called **piping**. This involves using a 
special symbol that looks like this `%>%`. This is equivalent to applying a function 
to an object. For example, to find the mean age, we could use the following code: 

```{r}
mean(my_data$age)
```

Alternatively, we could write the following code, using this special syntax: 

```{r}
my_data$age %>% 
  mean()
```

This line of code says to take the `df$age` column and apply the `mean()` 
function to it. Functionally, this is the same as doing `mean(df$age)`. 
We'll use this syntax when manipulating datasets. 

## Selecting specific columns

Suppose that we want to reduce
the dataset `df` to a specific set of columns - id, tx, age, and sex. We can use 
the `select()` function in the `dplyr` package:

```{r}
# select id, tx, age, and sex
# then, view the first few rows using head()
my_data %>% 
  select(id, tx, age, sex)
```

Note that this will NOT change the original `df` object. If we look at `df`, we can
still see that it has all of the original columns: 

```{r}
my_data
```

If we want to save the dataframe with the select columns, we can save it as a new 
object: 

```{r}
# create new variable df_select_columns
# which is a dataframe with only the id, tx, age, and sex columns
df_select_columns <- my_data %>% 
  select(id, tx, age, sex)

df_select_columns 
```

## Selecting specific rows

Next, we will learn how to select specific rows of a dataframe. We can filter 
using the `filter()` function based on some logic. For example, suppose that we 
want to create a dataframe which only includes BUP-NX (where tx takes the value 
of 0) participants: 

```{r}
my_data %>% 
  filter(tx == 0) 
```

Inside the `filter()` function, we specify the logic. In this example, you can see
that we specify `tx == 0`, which means that we only want to include treatment = 0 
participants in our filtered dataset. 

Note that we use a double equal sign `==`, NOT a single equal sign `=` in this 
logic statement. 

Some other examples: 

1. Greater or less than (< or >): 

```{r}
# filter for age > 30 participants 
my_data %>% 
  filter(age > 30) 
```

2. Not (using !)

```{r}
# filter for sex NOT equal to 1  
# use != to say "not equal to 
my_data %>% 
  filter(sex != 1) 
```

3. Composite logic using and (&) / or (|)

```{r}
# filter for male and BUP-NX participants 
my_data %>% 
  filter(sex == 0 & tx == 0) 
```

```{r}
# filter for male OR BUP-NX participants
my_data %>% 
  filter(sex == 0 | tx == 0) 
```

4. Missing rows using (`is.na()`)

```{r}
# select rows that are missing their mhdx value
my_data %>% 
  filter(is.na(mhdx))
```

```{r}
# select rows that are NOT missing their mhdx value
my_data %>% 
  filter(!is.na(mhdx)) 
```

**Example:** Explain what these lines of code do. Run the code and examine the 
`df_new` object to check your hypothesis.

```{r}
df_new <- my_data %>% 
  filter(!is.na(mhdx) & age > 30 & tx == 0) %>% 
  select(id, tx, age, mhdx)
```

**Example:** Write code to filter a dataframe for those 30 or older, select 
the `mhdx` column, and count the number of 0's and 1's in that column. How does it
compare to the number of 0's and 1's in the `mhdx` column for those under the
age of 30. 

```{r}
# write code below 

```

## Changing / creating new columns

We can also create / change new columns using `dplyr` with the `mutate()` 
function. The syntax for that looks like this:

* Start with the dataset and add a pipe `%>%` symbol. 
* Operate on the dataset with the `mutate()` 
* Within the `mutate()` function, specify `col = val`, where col is the 
column name that you want to add/change and val is what we want the column to 
look like (can reference other columns).

Some examples: 

```{r}
# creates a new column called age_months which is equal to the age column * 12 
my_data %>% 
  mutate(age_months = age * 12) %>% 
  select(id, age, age_months) %>% 
  head() 

```

```{r}
# replaces the id column with a vector from 1:100 
my_data %>% 
  mutate(id = 1:100) %>% 
  head() 

```

```{r}
# uses the cars R dataset, calculates time as distance divided by speed
cars %>% 
  mutate(time = dist / speed) %>% 
  head() 
```

You can also manipulate multiple columns as the same time; just list the things
you want to change sequentially inside the `mutate()` function, separated by 
commas. 

```{r}
# replaces the id column with a vector from 1:100 
my_data %>% 
  mutate(id = 1:100, 
         age_months = age * 12, 
         age_days = age_months * 12) %>% 
  head() 
```

Remember, just running these commands does not save the dataset in the variable. 
To do that, we will need to assign the object to the variable (or replace it with 
a new variable).

```{r}
cars_new <- cars %>% 
  mutate(time = dist/speed)

colnames(cars_new)
```

# Summaries for subgroups 

Remember how we can obtain descriptive summaries for the entire dataset, by 
using functions like `mean()`, `sd()`, `median()`, `min()`, and `max()`.

```{r}
mean(my_data$age)
median(my_data$bmi, na.rm = T)
```

What if we wanted to obtain these summaries for specific groups? Say, for example, 
we wanted to find the median `age`  for each treatment (`tx`). 

One option is to do this in two steps: 
* Create a subsetted dataset for each treatment 
* Manually find the median 

```{r}
my_data_tx0 <- my_data %>% filter(tx == 0)
my_data_tx1 <- my_data %>% filter(tx == 1)

median(my_data_tx0$age, na.rm = T)
median(my_data_tx1$age, na.rm = T)
```

With the `dplyr` package, you can do this all in one command: 
* The first step is to specify what groups you want to examine and place it as 
an argument in the `group_by()` function. 
* The next step is to configure what summaries you want to perform, and specify 
it in the `summary()` function. The syntax inside summary, for some generic summary
function, looks like `summary(label = function(column))`. For example:

```{r}
my_data %>% 
  group_by(tx) %>% # study each group
  summarise(mean_age = mean(age)) # want to pull the mean age for each treatment
```

**Example:** Use `group_by()` and `summarise()` to find the median BMI for each sex. 

```{r}
# write code below 

```

With `summarise()`, you can obtain multiple summaries at once. For example, suppose
that we wanted to get the mean age, standard deviation of the age, and the minimum 
age for each treatment group. 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(mean_age = mean(age), 
            sd_age = sd(age), 
            min_age = min(age)) 
```

You can also count the number of observations in each group with `n()`: 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(num_participants = n())
```
And, you can count the number of observations based on some logic using `sum()`, 
with a logic statement inside the parenthesis: 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(num_males = sum(sex == 0)) # count the number of males in each treatment arm
```

**Example:** Explain what the code below does. Modify it to add the proportion
of females in each treatment arm into the table. 

```{r}
# write code below 
my_data %>% 
  group_by(tx) %>% 
  summarise(num_participants = n(), 
            num_males = sum(sex == 0)) %>% 
  mutate(prop_males = num_males / num_participants)

```


# Exercises

1. Explain what these lines of code do. Run the code to check your answer. 

```{r}
my_data %>% 
  filter(sex == 0) %>% 
  nrow()
```

2. Explain what these lines of code do. Run the code to check your answer.

```{r}
df_new2 <- my_data %>% 
  filter(sex == 0)

summary(df_new2$age)

```

3. Explain what these lines of code do. Run the code to check your answer. 

```{r}
my_data %>% 
  filter(sex == 0) %>% 
  select(smoke100) %>% 
  table()
```

4. How old are the participants who are missing a value for `bmi`? Are they 
male or female? 

```{r}
# write code here 
```

5. Write code to create a new dataset called `more_educ_df` which first creates a
new column called `educ_months` (equal to `educyrs` multiplied by 12) and then
filters for the rows with more than 125 months of education. What proportion of 
participants in this new dataset is insured? 

```{r}
# write code here 
```

6. Write code to calculate the mean and standard deviation of BMI of those who 
are insured (`insure == 1`) versus those not who are not. 

```{r}
# write code here (HINT: use group_by() and summarise())
```

7. Hypothesize what this line of code does. Run the code and check the output to 
test your hypothesis. 

```{r}
modified_data <- my_data %>% 
  mutate(sex_str = ifelse(sex == 1, "Female", "Male")) %>% 
  select(id, sex, sex_str)
```

8. Use what you learned from exercise (7) to create new columns for `insure` and 
`smoke100` which have readable names for the cell values (e.g. yes / no) rather
than 0's and 1's.

```{r}
# write code here 
```
