---
title: 'CRISP Day 7 2026: Logistic regression'
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r, warning = F, message = F}
library(dplyr)
```

In today's lesson, we will learn how to adjust for covariates in linear regression 
as well as run and interpret logistic regression.  

```{r}
# read data 
my_data <- read.csv("crisp-2024-sample100.csv")
crisp_survey_data <- read.csv("crisp_survey_data_v2.csv")

# create factor variables 
my_data <- my_data %>% 
  mutate(sex = factor(sex, labels = c("male", "female")))

crisp_survey_data <- crisp_survey_data %>% 
  mutate(sex = factor(sex, labels = c("female", "male")), 
         age_group = factor(age_group, labels = c("under33", "over33")),
         pet_preference = factor(pet_preference, labels = c("dogs", "cats", "both", "neither")), 
         morning_or_evening_person = factor(morning_or_evening_person, labels = c("morning", "evening", "neither")), 
         interacted_with_clubs = factor(interacted_with_clubs, labels = c("no", "yes")), 
         spent_time_in_nature = factor(spent_time_in_nature, labels = c("no", "yes")))
```

# Linear regression with more than one predictor

Suppose that we run a linear regression, with number of years of education `educyrs` 
as the outcome, and `age` and `sex` as the predictors. That is, we are trying 
find the best estimates of $\beta_0$, $\beta_1$, and $\beta_2$ in the equation below: 

$$
educyrs = \beta_0 + \beta_1 age + \beta_2 sex
$$

To do this in R, we can use `lm()` with `age + sex` as our predictors in the formula: 

```{r}
linear_mod <- lm(educyrs ~ age + sex, data = my_data)

summary(linear_mod)

confint(linear_mod)
```

How can we interpret the output? 

* *Intercept*: The predicted number of years of education for males (sex = 0) with
age = 0. Note that this is not meaningful, since we don't have age = 0 in our
population of interest. 
* *Coefficient on age*: The estimated difference in number of years of education comparing 
two groups *of the same sex* differing by one year of age. 
* *Coefficient on sex*: The estimated difference in the number of years of education 
comparing females versus males *of the same age*. 

We can also pull the coefficients intervals all in one line using the `tidy()` 
function in the `broom` package. For this function, we should specify `conf.int = TRUE`. 

```{r}
# install.packages("broom")
library(broom)

tidy(linear_mod, conf.int = TRUE)

```


*Example*: In the CRISP survey data, regress anxiety/depression/irritability score (`anxious_depressed_or_irritable`) on hours of sleep (`hours_of_sleep`) 
adjusting for sex (`sex`). Interpret your result. 

```{r}
# write code here 

```


# Logistic regression with one predictor 

In logistic regression, the outcome is binary - either 0 or 1. For example, 
in this example, suppose that we consider history of psychotic diagnosis (`mhdx`) 
as our outcome, and age as our exposure. 

In the logistic model, we model the **odds** of the outcome. In this example, we 
are modeling the odds of having a history of psychotic diagnosis, which we
denote as $Odds(mhdx)$. Logistic regression also uses a logarithmic transform on 
this outcome, for reasons not explained in this tutorial. So our model is:  

$$
log(Odds(mhdx)) = \beta_0 + \beta_1 age
$$

The right hand side of the model should be familiar; this is a just a normal linear
model. 

Equivalently, we can exponentiate both sides and write this as: 

$$
Odds(mhdx) = exp(\beta_0 + \beta_1 age) 
$$

To run this in R, we can use the `glm()` function. `glm` stands for "generalized 
linear model," which is a more flexible form of a linear model. Within the `glm()` 
function call, we specify the: 

1. Formula: `outcome ~ predictor` (similar to linear regression)
2. Data: `data = df` 
3. Family: `family = binomial` (this configures the function to run logistic regression)

**Note**: The outcome can either be coded as a numeric (0/1) variable or as
a factor variable with two levels. 

So our function call looks like this: 

```{r}
mod1 <- glm(mhdx ~ age, data = my_data, family = binomial)

summary(mod1)
```
The estimates for the intercepts and age correspond to estimates for $\beta_0$ and
$\beta_1$, respectively. In logistic regression, it is often more useful to interpret 
the exponentiated coefficient estimates. To do that, we exponentiate the 
coefficient table using the `exp()` function on the `coef()` table from 
the `mod` object: 

```{r}
exp(coef(mod1))
```

How can we interpret these coefficients?

* *Intercept*: The exponentiated intercept represents the predicted odds of having a 
history of psychotic diagnosis for someone with age = 0. That is, we estimate that 
the odds of having a history of psychotic diagnosis for someone with age = 0 is 0.75. 
* *Coefficient on age*: The exponentiated coefficient on age represents the odds ratio 
comparing the odds of having a history of psychotic diagnosis for two groups differing 
by one year of age. That is, we estimate that for two groups differing in one year of age, 
the older group has around a 4% lower odds ($1 - 0.962$) of having a history of psychotic
diagnosis compared to the younger group. 
  * *p-value*: The result from testing the null hypothesis that the coefficient on 
  age is 0 (or equivalently the exponentiated coefficient is 1). With $p = 0.275$, we 
  fail to find a significant association between age and history of psychotic diagnosis. 

To obtain confidence intervals for these exponentiated estimates, we can 
exponentiate the output of `confint(mod)`: 

```{r, message = F}
exp(confint(mod1))
```

We can also pull the exponentiated coefficients and confidence intervals all 
in one line using the `tidy()` function in the `broom` package. For this function, 
we should specify `exponentiate = TRUE` (for logistic regression) and `conf.int = TRUE`. 

```{r}
# install.packages("broom")
library(broom)

tidy(mod1, exponentiate = TRUE, conf.int = TRUE)
```
**Example**: In the CRISP survey data, run a logistic regression with
`spent_time_in_nature` as the binary outcome and `hours_of_sleep` as the 
predictor. Interpret your result. 

```{r}
# write code here
```

# Logistic regression with more than one predictor 

Suppose that we run a logistic regression with history of psychotic diagnosis (`mhdx`)
as the outcome, and `tx` and `age` as the predictors. That is, we are trying 
find the best estimates of $\beta_0$, $\beta_1$, and $\beta_2$ in the equation below: 

$$
log(Odds(mhdx)) = \beta_0 + \beta_1 treatment +  \beta_2 age
$$

Equivalently, we can exponentiate both sides and write this as: 

$$
Odds(mhdx) = exp(\beta_0 + \beta_1 treatment + \beta_2 age) 
$$

Like in the previous section, we run this in R using the `glm()` function. In this case, 
we specify the formula as `mhdx ~ tx + age`, with `mhdx` as the outcome, and 
`tx` and `age` as the predictors 

```{r}
mod2 <- glm(mhdx ~ tx + age, data = my_data, family = binomial)

summary(mod2)
```

Again, we interpret the exponentiated coefficients (as well as their confidence
intervals). 

```{r}
exp(coef(mod2))
exp(confint(mod2))

# alternatively use tidy() in the broom package
tidy(mod2, exponentiate = TRUE, conf.int = TRUE)
```
* *Intercept*: The exponentiated intercept represents the predicted odds of having a 
history of psychotic diagnosis for someone with age = 0 and tx = 0. That is, we estimate that 
the odds of having a history of psychotic diagnosis for someone with age = 0 and 
tx = 0 is 0.64. 
* *Coefficient on tx*: The exponentiated coefficient on tx represents the odds ratio 
comparing the odds of having a history of psychotic diagnosis for those with tx = 1 
to that for those with tx = 0 of the same age. That is, comparing individuals of the same 
age, we estimate that those with tx = 1 have a 48% higher odds ($1.48 - 1$) of 
psychotic diagnosis than those with tx = 0.
  * *p-value*:  The result from testing the null hypothesis that the coefficient on 
  tx is 0 (or equivalently the exponentiated coefficient is 1). With $p = 0.459$, we 
  fail to find a significant association between tx and history of psychotic diagnosis, 
  after controlling for age. 
* *Coefficient on age*: The exponentiated coefficient on age represents the odds ratio 
comparing the odds of having a history of psychotic diagnosis for two groups differing 
by one year of age with the same treatment status. That is, we estimate that for two groups
with the same treatment status differing in one year of age, the older group has 
around a 4% lower odds ($1 - 0.962$) of having a history of psychotic diagnosis 
compared to the younger group. 
  * *p-value*: The result from testing the null hypothesis that the coefficient on 
  age is 0 (or equivalently the exponentiated coefficient is 1). With $p = 0.274$, we 
  fail to find a significant association between age and history of psychotic diagnosis
  after controlling for treatment status.
  
**Example**: In the CRISP survey data, run a logistic regression with 
`interacted_with_clubs` as the binary outcome and `age_group` as the predictor, 
adjusting for `sex`. Interpret your result. 

```{r}
# write code here
```

# Exercises 

For the exercise below, use the crisp-2024-sample100.csv dataset (`my_data`). 

1. Perform logistic regression with cannabis use (`canlft`) as the outcome and 
sex (`sex`) as the predictor. Obtain estimates and confidence intervals for the
exponentiated coefficients. Provide an interpretation of the output. 

```{r}
# write code here
```

2. Perform logistic regression with cannabis use (`canlft`) as the outcome and 
sex (`sex`) and treatment status (`tx`) as predictors. Obtain estimates and confidence
intervals for the exponentiated coefficients. Provide an interpretation 
of the output. 

```{r}
# write code here
```


