---
title: 'CRISP 2026: Homework 3'
output:
  html_document:
    toc: yes
    toc_float: yes
---

# Part 1: Choose your research question 

In this final homework, we will conduct a guided analysis of the CRISP survey 
data. The goal of this homework is to apply everything that you learned this 
quarter for a real analysis. 

**Guidelines for the analysis**: 

* The study population should exclude instructors of CRISP.

* All analyses should adjust for age group. 

* Use the dataset specifically created for this homework: crisp_survey_data_hw3.csv.
The data dictionary is on the course website. 

**Choose your research question**: 

1. Choose two outcomes (one in each column) and one exposure of interest that 
you wish to study.

| Outcome 1 (numeric)      | Outcome 2 (binary)                                   | Exposure                         |
|--------------------------|------------------------------------------------------|----------------------------------|
| Cardio exercise          | Able to deal with stressors                          | Spent time in nature             |
| Stress / anxiety score   | Woke up refreshed                                    | Morning or evening person        |
| Pain score               | Satisfied with social activities and relationships   | Prefer home cooking or not       |
| Sleep quality            |                                                      | Drink alcohol                    |

* Outcome 1: 

* Outcome 2: 

* Exposure: 


2. Write your *research question of interest*. There should be two research questions 
of interest, one for each outcome. An example is: We aim to examine the association between 
weekly minutes of cardio exercise and age, adjusted by sex, among CRISP students
and their friends.

* Research Question 1: 

* Research Question 2: 


3. Write your *analysis plan*. Specify how you wish to analyze the data. 
An example is: We plan to run a linear regression with cardio exercise as the outcome, 
age as the predictor, and sex as the adjustment variable. 

* Analysis plan for Research Question 1: 

* Analysis plan for Research Question 2: 



# Part 2: Conduct your analysis 

1. Read in the CRISP survey data file (crisp_survey_data_hw3.csv) used for this
homework assignment. 

```{r}
# write code here 

```

2. Perform cleaning/processing of your dataset. You may wish to convert some 
variables to factor variables and/or filter your dataset for select rows. 

```{r}
# write code here

```

3. Create a Table 1 describing characteristics, stratified by your exposure of 
interest. Include at least five variables in your Table 1 but do not include your
outcomes of interest. 

```{r}
# write code here 

```

4. Are there any significant differences in characteristics between the
two groups as described in the Table 1 above? 

```{r}
# write code here 

```

5. Conduct an analysis to answer your first research question of interest. 
Interpret your result! (Don't just give R output.)

```{r}
# write code here 

```

6. Conduct an analysis to answer your second research question of interest. 
Interpret your result! (Don't just give R output.)

```{r}
# write code here 

```

7. Bonus: Create a graph to display your result. 

```{r}
# write code here 

```

