---
title: 'CRISP 2026: Homework 1'
output: html_document
---

1. Load in the library `dplyr` into R. 

```{r}
# write code here 
library(dplyr)
```

2. Download in the `hw1_fake_data.csv` dataset, load it into R, and save it as 
the variable `df`. 

```{r}
# write code here 
df <- read.csv("hw1_fake_data.csv")
```

Here is the data dictionary for this dataset:

* `id`: Unique person identifier.
* `treatment`: Treatment group indicator; `0` = control, `1` = treatment.
* `gender`: Participant gender; `male` or `female`.
* `weight`: Participant weight in pounds. Some values are missing.
* `diabetes`: Diabetes status indicator; `0` = no diabetes, `1` = has diabetes.

3. Count the number of rows in the dataset. 

```{r}
# write code here 
nrow(df)
```

4. How many females are in the dataset? How many participants are on treatment? 
How many participants have diabetes? 

```{r}
# write code here 
table(df$gender)
table(df$treatment)
table(df$diabetes)
```

5. Provide summary statistics for the weights of all participants. How many 
participants are missing the `weight` variable? 

```{r}
# write code here 
summary(df$weight)
```

6. Subset the dataset into participants who have diabetes and participants who 
don't have diabetes, selecting only the columns `id` and `diabetes`. Assign these 
into variables with names `diabetes_df` and and `no_diabetes_df`, respectively. 

```{r}
# write code here 
diabetes_df <- df %>% 
  filter(diabetes == 1) %>% 
  select(id, diabetes)

no_diabetes_df <- df %>% 
  filter(diabetes == 0) %>% 
  select(id, diabetes)

diabetes_df
no_diabetes_df
```

7. Give summary statistics for participants' weights comparing those with 
diabetes versus those without. 

```{r}
# write code here 
df %>% 
  group_by(diabetes) %>% 
  summarise(mean_weight = mean(weight, na.rm = T), 
            sd_weight = sd(weight, na.rm = T), 
            min_weight = min(weight, na.rm = T), 
            max_weight = max(weight, na.rm = T))
```

8. Comparing treated versus untreated participants: 
* How many participants are in each group? 
* How many participants have diabetes in each group? 
* How many participants are male in each group? 
* What proportion in each group have diabetes? 
* What proportion in each group are male? 
(Hint: Use `group_by()`, `summarise()`, and `mutate()`)

```{r}
# write code here 
df %>% 
  group_by(treatment) %>% 
  summarise(num_participants = n(), 
            has_diabetes = sum(diabetes == 1), 
            male_count = sum(gender == "male")) %>% 
  mutate(prop_diabetes = has_diabetes / num_participants, 
         prop_male = male_count / num_participants)
```
