---
title: 'CRISP 2026: Homework 2'
output: html_document
---

The goal of this practicum is to apply what we've learned so far on a new 
dataset! 

The dataset is the **midterm_practicum_sample.csv** file on the course website. 
Download this dataset and save it in your CRISP R notes folder. 

The dataset shows data from two randomized studies (Study 1 and Study 2), which 
study the efficacy of a treatment on some disease's severity. 
The dataset has the following structure: 

* id: 1 to 1000 
* study: 1 = study 1, 2 = study 2
* tx: 0 = placebo, 1 = treatment 
* age: numeric age variable 
* alcohol_usage: 0 = no/low, 1 = medium, 2 = high 
* disease_severity: numeric disease severity variable 

The goal is to create a Table 1, with the results of hypothesis testing comparing
the two treatment arms. 

1. Read in the the dataset and save it as the `df_hw` variable. Load in the 
`dplyr` and `table1` package. 

```{r}
# write code below 

```

2. You wish to analyze data for Study 1 ONLY. Also, you would like to remove the 
children (age < 18) from the dataset. Also, you'd like to create a new variable 
called `disease_severity_scaled` which equals `disease_severity` divided by 100.
Create a new dataset called `df_filter` which includes only the rows of interest 
and creates this new variable. 

```{r}
# write code below
```

3. Convert the `tx` and `alcohol_usage` to factor variables. Re-save it the new
dataframe into your `df_filter` variable. 

```{r}
# write code below 
```

4. Use the `table1` package to create a Table 1, summarizing age, alcohol usage, 
and disease severity stratified by treatment status. 

```{r}
# write code below 
```

5. Perform hypothesis tests to test the differences in these variables between
the treatment and placebo arms. What is the null hypothesis for each hypothesis
test? Provide an interpretation for the results. 

```{r}
# write code below
```
