---
title: 'CRISP Day 4 2026: dplyr and table1'
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
library(dplyr)
```

In today's lesson, we will learn about how to group and summarize data using
the `dplyr` and `table1` package. 

```{r}
my_data <- read.csv("crisp-2024-sample100.csv")
```

# Review exercise 

**Example:** Create a new column called `risk_score` which adds up the columns 
`canlft`, `smoke100`, and `coclft`. 

What is the mean score among men (`sex == 0`)? What is the mean score among women
(`sex == 1`)?

Hint: First, use `mutate()` to create the new variable. Then use `filter()`
to filter for the subgroups. Remember that you can create and reference new 
variables as you go along. 

```{r}
# write code here

```

# Summaries for subgroups (copied from Day 3)

Remember how we can obtain descriptive summaries for the entire dataset, by 
using functions like `mean()`, `sd()`, `median()`, `min()`, and `max()`.

```{r}
mean(my_data$age)
median(my_data$bmi, na.rm = T)
```

What if we wanted to obtain these summaries for specific groups? Say, for example, 
we wanted to find the median `age`  for each treatment (`tx`). 

One option is to do this in two steps: 
* Create a subsetted dataset for each treatment 
* Manually find the median 

```{r}
my_data_tx0 <- my_data %>% filter(tx == 0)
my_data_tx1 <- my_data %>% filter(tx == 1)

median(my_data_tx0$age, na.rm = T)
median(my_data_tx1$age, na.rm = T)
```

With the `dplyr` package, you can do this all in one command: 
* The first step is to specify what groups you want to examine and place it as 
an argument in the `group_by()` function. 
* The next step is to configure what summaries you want to perform, and specify 
it in the `summarise()` function. The syntax inside summary, for some generic summary
function, looks like `summarise(label = function(column))`. For example:

```{r}
my_data %>% 
  group_by(tx) %>% # study each group
  summarise(mean_age = mean(age)) # want to pull the mean age for each treatment
```

**Example:** Use `group_by()` and `summarise()` to find the median BMI for each sex. 

```{r}
# write code below 

```

With `summarise()`, you can obtain multiple summaries at once. For example, suppose
that we wanted to get the mean age, standard deviation of the age, and the minimum 
age for each treatment group. 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(mean_age = mean(age), 
            sd_age = sd(age), 
            min_age = min(age)) 
```

You can also count the number of observations in each group with `n()`: 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(num_participants = n())
```

And, you can count the number of observations based on some logic using `sum()`, 
with a logic statement inside the parenthesis: 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(num_males = sum(sex == 0), # count the number of males in each treatment arm
            missing_mhdx = sum(is.na(mhdx))) # count the number missing mhdx
```

**Example:** Explain what the code below does. Modify it to add the proportion
of females in each treatment arm into the table. 

```{r}
# write code below 
my_data %>% 
  group_by(tx) %>% 
  summarise(num_participants = n(), 
            num_males = sum(sex == 0)) %>% 
  mutate(prop_males = num_males / num_participants)

```

# CRISP survey data exploration

Let's explore the CRISP survey data using the tools that you have learned so far! 

* `record_id`: Unique REDCap survey record number.
* `age_group`: Age category (`0` = 33 or younger; `1` = older than 33).
* `sex`: Sex (`1` = Woman; `2` = Man; `3` = Non-binary; `4` = Prefer not to say).
* `pet_preference`: Pet preference (`1` = Dogs; `2` = Cats; `3` = Both; `4` = Neither).
* `morning_or_evening_person`: Time-of-day preference (`1` = Morning; `2` = Evening; `3` = Equal or neither).
* `food_preference`: Food preference (`1` = Gourmet; `2` = Upscale; `3` = Home cooking; `4` = Fast food).
* `spent_time_in_nature`: Spent at least two hours in nature during the past week (`0` = No; `1` = Yes).
* `interacted_with_clubs`: Interacted with at least one club or organization during the past week (`0` = No; `1` = Yes).
* `resistance_training_workouts`: Number of resistance-training workouts during the past week (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `hours_of_sleep`: Average hours of sleep per night (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `packaged_snacks_per_day`: Average number of packaged snacks consumed per day (`0` = 0 or 1; `2`–`9` = corresponding number; `10` = 10 or more).
* `anxious_depressed_or_irritable`: Frequency of feeling anxious, depressed, or irritable (`1` = Never; `2` = Rarely; `3` = Sometimes; `4` = Often; `5` = Always).
* `fatigue_rating`: Average fatigue rating (`1` = None; `2` = Mild; `3` = Moderate; `4` = Severe; `5` = Very severe).

```{r}
crisp_survey_df <- read.csv("crisp_survey_data_v1.csv")
```

One example: find the mean number of hours of sleep that those 33 or younger get? 
What about over 33? 

```{r}
crisp_survey_df %>% 
  group_by(age_group) %>% 
  summarise(count = n(), 
            mean_sleep = mean(hours_of_sleep, na.rm = T))
```

Come up with your own questions and explore the data: 

```{r}
# write code here 
```


# Creating Table 1's 

We have learned everything that we need to make a "Table 1" by hand. 

```{r}
my_data %>% 
  group_by(tx) %>% 
  summarise(mean_age = mean(age), 
            sd_age = sd(age))
```

The problem with this can be a very manual, time-intensive process, if we 
have to do this for every variable. Alternatively, we can use the `table1` 
package to help us do this. 

## Factor variables 

Before we get into learning how to use the `table1` package, we will need to 
first learn about **factor** variable types. 

Each dataframe column can have different types. Some variable are *string* variables 
(e.g. text); others are *numeric* variables (e.g. numbers). There is another type 
of variable in R, called a **factor**. These are categorical variables, which often 
are coded as integers in a dataset (e.g. 0, 1, 2, etc.). 

One example: we may want to consider the variable `tx` in the crisp-2024-sample100.csv 
dataset as a factor variable because it only takes two distinct values: 0 and 1, where
0 corresponds to BUP-NX and 1 corresponds to XR-NTX. 

```{r}
table(my_data$tx)
```

By default, R will read in number variables as in *numeric*. It starts off by 
thinking this column is just a number. But, we may want to change it into a 
factor before we perform our analysis. To do this, we can use the following line of code: 

```{r}
# using base R 
# the labels correspond to the order of the numbers (i.e. 0 = BUP-NX, 1 = XR-NTX)
my_data$tx_factor <- factor(my_data$tx, labels = c("BUP-NX", "XR-NTX"))
my_data[ , c("tx", "tx_factor")]
```

```{r}
# using dplyr
my_data %>% 
  mutate(tx_factor = factor(tx, labels = c("BUP-NX", "XR-NTX"))) %>% 
  select(tx, tx_factor)
```

**Example:** Complete the following code to convert the following variables 
into factor variables with the following labels: 

* `sex`: 0 = males, 1 = females 
* `insure`: 0 = uninsured, 1 = insured
* `smoke100`: 0 = no, 1  = yes
* `mhdx`: 0 = no, 1 = yes

```{r}
# complete the "FILL IN HERE" blanks in the code below 
my_data_factor <- my_data %>% 
  mutate(tx = factor(tx, labels = c("BUP-NX", "XR-NTX")),
         sex = "FILL IN HERE", 
         insure = "FILL IN HERE", 
         smoke100 = "FILL IN HERE", 
         mhdx = "FILL IN HERE") 

```

**Note**: Remember the `summary()` function? Once we convert the variables into 
factors, we can now use it to correctly summarize categorical variables in the data. 

```{r}
summary(my_data_factor)
```

## Using `table1` package

Let's read in the package: 

```{r, message = F}
library(table1)
```

Before we use this package, we will need to convert our categorical variables 
into factors. Luckily, we already did this in Example 3. This is a necessary 
first step if there are coded categorical variables, or else the `table1` 
package will this that these columns are numeric!

After this, we can run the `table1()` function: 

```{r}
table1(~ age | tx, data = my_data_factor)
```

The `table1()` function includes two argument: 

* First, we specify the formula for
how we want the table to look. This formula takes the form `~ var1 | var2`, where
`var1` are the columns that we want to summarize, followed by a `|`, and then `var2`, 
which is the column that we want to stratify by. So `~ age | tx` means that we want to 
summarize the age column, stratified across the two treatment arms. 
* Next, we specify the dataframe, which is why we include a `data = my_data_factor` additional argument. 

We can add variables to summarize by using `+`. The following line of code summarizes 
both the age and sex variable across treatment arms: 

```{r}
table1(~ age + sex | tx, data = my_data_factor)
```
Note that by default, `table1` will provide the mean, median, min, max, and standard 
deviation for continuous numeric variables and proportions for categorical factor variables.

We can change the labels in these Table 1's by setting labels 
for each column of interest. To do this, the code looks like this: 

```{r}
label(my_data_factor$sex) <- "Sex"
label(my_data_factor$age) <- "Age"

table1(~ age + sex | tx, data = my_data_factor)

```


# Exercises

1. Write code to calculate the mean and standard deviation of BMI of those who 
are insured (`insure == 1`) versus those not who are not. 

```{r}
# write code here (HINT: use group_by() and summarise())
```

2. Write code to show descriptive statistics of the CRISP survey data using 
the `table1` package. 

```{r}
# write code here 
```

