---
title: 'CRISP Day 1 2026: Getting starting with R'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
---

In today's lesson, we will get our computer set up with R. 

# Installing R 

If you have not yet, please start by downloading and installing R and R Studio. 

1. Download R ([link](https://cran.r-project.org/)) and run the installation file. 
2. Download R Studio ([link](https://posit.co/download/rstudio-desktop/)) and run the installation file. 

# Folder setup 

1. Please create a folder somewhere on your computer called **crisp_R_notes**
(or something similar). 
2. Within this folder, please download and save the data CSV file from the 
[course website](https://jpspeng.github.io/crisp_notes/).
3. Within this folder, please download and save the R Markdown notebook (.Rmd) for today 
(July 6). For future lessons, please put all notebooks within this folder. 

# Creating, saving, and opening R files  

1. Open R Studio. Click on File -> New File -> R Script. Then, press File -> Save As, 
and save the file as *test.R* in your crisp_R_notes folder. 
2. Click on File -> New File -> R Notebook. Then, press File -> Save As, 
and save the file as *test.Rmd* in your crisp_R_notes folder. 
3. Open today's R Markdown file (*crisp_2026_day1.Rmd*) and continue this 
tutorial within R Studio. 


# Code chunks and comments 

Code within R notebooks run within code chunks. To add a code chunk, you can press
the +C icon on the right hand corner of the Source window, and click on R.

To run the code chunk, **press the play button** on the upper right hand corner of the
code chunk: 

```{r}
5 + 5
```

Code that has a # before it is a comment and will not run. Comments are often 
used for documentation - so that you remember what code does after you write it. 

Try running this line of code, and see what happens: 

```{r}
# 5 + 5 
```


# Installing and using packages 

One of the main advantages of R are its packages. 

An R package is a collection of functions, data, and documentation that extends 
the capabilities of R. They are used to share code and data with others and to 
reuse code in different projects. They can include:

1. **Functions**: Custom functions that perform specific tasks.
2. **Data**: Datasets that can be used for analysis and examples.
3. **Documentation**: Help files and vignettes that explain how to use the functions 
and datasets.

R packages are typically installed from CRAN (Comprehensive R Archive Network). Packages
tend to have specific uses. For example, there are packages specifically for 
making graphs, others for creating manuscript-ready tables, and others to simplify 
data processing. 

You can install a package in R using the `install.packages()` function, and load 
it using `library()`.

```{r}
# Install a package from CRAN
install.packages("dplyr")
```

```{r}
# Load the package
library(dplyr)
```

# Reading in data into R 

Say that you have a dataset. The first step is to read in data into R. You can do
this using the `read.csv()` function. 

The easiest way to use this function is to put the data file in the **same folder** 
as your R (or Rmd) code. You should have done this already with the folder setup 
above. Try to run the following code: 

```{r}
read.csv("crisp-2024-sample100.csv") 
```

Did it run? Then great! That means that the data is now in our R session. We can 
now manipulate and analyze the data, which we will do in the next few tutorials.

If the data is not in the same folder as your R file, you may have to include the
directory path to the data in the argument when calling `read.csv()`.

```{r}
# if data is in a different folder, it may look like this: 
# df <- read.csv("C:/path/to/directory/crisp-2024-sample100.csv")
```

# Expressions 

An **expression** is a line of code that just evaluates. Nothing is saved - the output 
is simply evaluated and printed. For example, we can perform the operation 5 + 5 and 
print the result in R. 

```{r}
5 + 5
```

Once we complete an expression, we can write a new line of code in the next line. 
The two lines of code will run sequentially. 

```{r}
5 + 5
5 + 10
```

If we do not complete an expression, the next line will run as a continuation of the 
previous line. 

```{r}
5 + 5 + 
  10 
```

# Variables 

We can save **expressions** into **variables**. We can do this using the = or 
<- symbol. (The <- symbol is more standard in R.) 

For example, the line of code below saves 5 + 5 into a new variable `my_num`: 

```{r}
my_var <- 5 + 5 

my_var
```

After we save a variable, we can reference and perform operations on this variable 
later on in our code: 

```{r}
my_var + 10
```

Note that doing this operation does not change our variable. If we take a look at 
`my_var`, it still equals 10. 

```{r}
my_var
```

We can change our variable by assigning the result of `my_var + 10` to `my_var`. 

```{r}
my_var <- my_var + 10 

my_var
```


A few more things about variables: 

* Variable names cannot contain spaces. 
* Try to be descriptive with your variable names!  For example, if you save an 
variable which holds the number of participants in a study, it is better to name
it something like `num_participants` instead of something generic, like `a` or `b`.
* R is cap-sensitive. If we save a variable `MY_NUM <- 5`, and try to reference 
lowercase `my_num` later on, it will cause an error. 

**Example**: Create a new variable called `test_var`. Set it equal to 10. 
Then create another variable called `new_test_var` and set it equal to `test_var` 
multiplied by 20. Print `new_test_var`. 

```{r}
# write code here 

```

# Variable types 

We can save different types of objects into variables. Three of the most common 
types of objects are **numeric** (numbers), **string** (text), and **boolean** 
(TRUE/FALSE) variables.

```{r}
my_num <- 5 

my_str <- "This is my string"

# Booleans are TRUE/FALSE 
# They can also be written as T/F 
my_boolean <- TRUE
```

# Vectors

A **vector** is an ordered collection of numeric or string objects. We can create
a vector by using the `c()` syntax. For example, the code below creates a vector of 
5 numbers, from 1 to 5. 

```{r}
my_vec <- c(1, 3, 5, 7, 9)
```

We can select specific elements of the vector by calling the vector, followed by 
`[]`. For example, this line of code selects the first element:  

```{r}
my_vec[1]
```

This line of code selects the fourth element of the vector: 

```{r}
my_vec[4]
```

This line of code selects the first and fourth element of the vector: 

```{r}
my_vec[c(1,4)]
```

This line of code selects the first four elements of the vector. 

```{r}
my_vec[1:4]
```

(Note: the `a:b` syntax is a shortcut for writing a vector `c(a, a+1,...,b-1, b)`)

Vectors can also consist of strings. 

```{r}
my_str_vec <- c("a", "b", "c")

my_str_vec[1]
```

R operations are **vectorized**, meaning that you can apply an operation onto a 
vector, and it will do the operation across every element of the vector. 

```{r}
my_even_vec <- my_vec * 2

my_even_vec
```

**Example**: Create a new variable vector called `test_vec`. Set it equal to 
the a vector containing the numbers 1 through 5. Multiply the vector by 10, 
and then save the result as a new vector called `new_test_vec`. Then select 
the 2nd through 4th element of `new_test_vec`. 

```{r}
# write code here 

```

# Exercises 


1. Install the following packages: `ggplot2`, `table1`, and `survival`. 

```{r}
# type in code here
install.package("ggplot2")
install.package("table1")
install.package("survival")

```

2. Use the `+`, `-`, `*`, and `/` operators to:
* Add 1048 and 890
* Multiply the previous result by 20
* Subtract the previous result by 50 
* Divide the previous result by 2.5

```{r}
# type in code here 
1048 + 890
1938 * 20 
38760 - 50
38710 / 2.5
```

3. Describe what the following code is doing: 

**Answer**: The code is doing the same series of operations as Q2.  

```{r}
a <- 1048 
b <- a + 890
c <- b * 20 
d <- c - 50 
e <- d / 2.5 

print(e)
```

4. Following the pattern in the previous code block, perform the following operations: 
* Multiply 13 and 631
* Divide the previous result by 25 
* Add 10 to the previous result 

```{r}
# type in code here 
a <- 13
b <- a * 631 
c <- b / 25 
d <- c + 10 

d
```

5. Describe what the code below is doing: 

**Answer**: The code is adding 5 to each element of the list 1,2,3,4,5. 

```{r}
a <- c(1,2,3,4,5)
b <- a + 5

print(b)
```

5. Describe what the code below is doing: 

**Answer**: The code is adding 5 to each element of sequence 5,6,7,8,9,10. 

```{r}
a <- 5:10
b <- a + 5

print(b)
```

6. Use what you learned from the previous questions to multiply all numbers 
from 25 through 31 by 13. 

```{r}
# type in code here 
a <- 25:31 
b <- a * 13 

print(b)
```